---
output:
  github_document:
    toc: true
    toc_depth: 2
  html_document:
    toc: true
    theme: cerulean     # or bootstrap, etc.
    mathjax: default
editor_options: 
  markdown: 
    wrap: sentence
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%",
  echo = TRUE,
  message = FALSE,
  warning = FALSE
)
```

<!-- badges: start -->

[![Binder](https://mybinder.org/badge_logo.svg)](https://mybinder.org/v2/gh/nithecs-biomath/RBasicPack/master?urlpath=rstudio) [![Lifecycle: stable](https://img.shields.io/badge/lifecycle-stable-brightgreen.svg)](https://lifecycle.r-lib.org/articles/stages.html#stable) [![test-coverage](https://github.com/macSands/dissmapr/actions/workflows/test-coverage.yaml/badge.svg)](https://github.com/macSands/dissmapr/actions/workflows/test-coverage.yaml) [![Codecov test coverage](https://codecov.io/gh/macSands/dissmapr/graph/badge.svg)](https://app.codecov.io/gh/macSands/dissmapr) [![R-CMD-check](https://github.com/macSands/dissmapr/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/macSands/dissmapr/actions/workflows/R-CMD-check.yaml)

<!-- badges: end -->

# `dissmapr`

## A Novel Framework for Automated Compositional Dissimilarity and Biodiversity Turnover Analysis

## Introduction

`dissmapr` is an R package for analysing compositional dissimilarity and biodiversity turnover across spatial gradients.
It provides scalable, modular workflows that integrate species occurrence, environmental data, and multi-site metrics to quantify and predict biodiversity patterns.
A core feature is the use of zeta diversity, which extends beyond pairwise comparisons to capture shared species across multiple sites---offering deeper insight into community assembly, turnover, and connectivity.
By incorporating modern approaches such as multi-site Generalised Dissimilarity Modelling (MS-GDM), `dissmapr` enables robust mapping, bioregional classification, and scenario-based forecasting.
Designed for flexibility and reproducibility, it supports biodiversity monitoring and conservation planning at landscape to regional scales.

------------------------------------------------------------------------

## Step-by-Step Workflow

`dissmapr` implements a structured, reproducible workflow for analysing biodiversity patterns and delineating bioregions.
Each function aligns with a specific step, guiding users from data acquisition to predictive mapping.
The workflow begins with sourcing species occurrence and georeferenced environmental data, followed by data formatting and calculation of compositional turnover using zeta diversity metrics (via the `zetadiv` package).
Multi-Site Generalized Dissimilarity Modelling (MS-GDM) is then applied to model and predict dissimilarity across landscapes.
These predictions feed into the Dissimilarity Cube, which classifies spatial clusters of species composition into distinct bioregions.
The framework supports the integration of historical and future climate data to assess shifts in biodiversity and detect emerging, shifting, or dissolving bioregions under global change.
This step-by-step structure, mirrored in the accompanying tutorial sections, promotes accessibility, transparency, and ecological insight at multiple spatial and temporal scales.

------------------------------------------------------------------------

### 1. Install and load `dissmapr`

Install and load the `dissmapr` package from GitHub, ensuring all functions are available for use in the analysis workflow.

```{r install, eval=FALSE, include=TRUE}
# install remotes if needed
# install.packages("remotes")
remotes::install_github("macSands/dissmapr")
```

```{r load-dissmapr, include=TRUE}
# Ensure the package is loaded when knitting
library(dissmapr)

# Make sure all the functions are loaded
devtools::load_all()
```

------------------------------------------------------------------------

### 2. Load other R libraries

Load core libraries for spatial processing, biodiversity modelling, and visualization required across the `dissmapr` analysis pipeline.

```{r libraries}
# Load necessary libraries
library(httr)       # HTTP client  
library(geodata)    # Download geographic data  
library(data.table) # Fast large-table operations  
library(dplyr)      # Data manipulation verbs  
library(tidyr)      # Tidy data reshaping  
library(zoo)        # Time series utilities  
library(sf)         # Vector spatial data  
library(terra)      # Raster spatial operations  
library(zetadiv)    # Multi-site biodiversity turnover  
library(ggplot2)    # Grammar of graphics  
library(viridis)    # Perceptual color scales  
```

------------------------------------------------------------------------

### 3. User-defined area of interest and grid resolution

Load the spatial boundary data for South Africa to serve as the geographic reference for all subsequent biodiversity analyses and visualizations.

```{r aoi}
# Read RSA shape file
rsa = sf::st_read('inst/extdata/rsa.shp')

# Define your resolution and create mask to use later
res = 0.5 # 0.5 degrees is roughly 55km

# Convert to a terra vector
rsa_vect = vect(rsa)

# Create an empty raster over RSA at your desired resolution
grid = rast(rsa_vect, resolution = res, crs = crs(rsa_vect))
values(grid) = 1   # fill with dummy values
```

------------------------------------------------------------------------

### 4. Get species occurrence records using `get_occurrence_data`

This section focuses on automating the retrieval and pre-processing of biodiversity occurrence data from various sources, including:

i)  local `databases`.csv` files,
ii) URLs or `.zip` files from the Global Biodiversity Information Facility (GBIF), and
iii) species occurrence cubes from B3 (specification) [*work in progress*].

The function assembles data on species distributions across specified taxonomic groups and regions, producing presence-absence or abundance matrices that quantify species co-occurrence within locations.

```{r get-occurrence}
bfly_data = get_occurrence_data(
  data        = 'inst/extdata/gbif_butterflies.csv',
  source_type = 'local_csv',
  sep         = '\t'
)
dim(bfly_data)
```

------------------------------------------------------------------------

### 5. Format data using `format_df`

Use `format_df` to standardise and reshape biodiversity data into long or wide formats.
The function automatically identifies key columns (e.g., coordinates, species, and values), assigns missing site IDs, and reformats the data for analysis.
Outputs include a cleaned dataset and species-site matrices for further processing:

• site_xy: Holds spatial coordinates of sampled sites.\
• site_sp: Site-by-species matrix for biodiversity assessments.

#### Format data into long and wide formats

```{r format-df}
bfly_result = format_df(
  data        = bfly_data,
  species_col = 'verbatimScientificName',
  value_col   = 'pa',
  extra_cols  = NULL,
  format      = 'long'
)
site_obs = bfly_result$site_obs
site_spp = bfly_result$site_sp
dim(site_obs)
dim(site_spp)
```

#### Get parameters to use later

```{r spp-cols}
# Get parameters from processed data
# Number of species columns
(n_sp = dim(site_spp)[2] - 3)
sp_cols = names(site_spp)[-c(1:3)]
```

------------------------------------------------------------------------

### 6. Summarise records by grid centroid using `generate_grid`

Use `generate_grid` to divide the study area, derived from the geographic extent of the occurrence data above, into grids of user-defined resolution, creating a spatial grid over a specified geographic extent.
It assigns unique grid IDs to points and summarizes selected data columns within each grid cell.
The function outputs a raster grid, grid polygons for visualization, and a data frame summarizing the contents of each grid cell, including totals and centroids.
It is particularly useful for aggregating spatial data, such as biodiversity observations, into predefined spatial units for further analysis.

#### Assign records to a grid at a set resolution (e.g. 0.5°)

Aggregate species records into equal-area grid cells to enable spatially standardized biodiversity analyses.

```{r grid}
grid_list = generate_grid(
  data       = site_spp,
  x_col      = "x",
  y_col      = "y",
  grid_size  = 0.5,
  sum_col_range = 4:ncol(site_spp),
  crs_epsg   = 4326
)

aoi_grid = grid_list$grid_sf
grid_spp = grid_list$block_sp
dim(aoi_grid)
dim(grid_spp)
```

#### Generate a data frame 'xy' of site centroids

Extract longitude--latitude coordinates and summary metrics for each occupied grid cell.

```{r site-xy}
# Grid centroids with 'gird_id', 'centroid_lon', 'centroid_lat', 'obs_sum' and `spp_rich`
grid_xy = grid_spp[,c(1:3,5:6)]
dim(grid_xy)
spp_obs = site_obs
dim(spp_obs)
```

#### Generate a map of RSA with occupied grid cells as centroid points

Visualize observation density across South Africa using centroid-based mapping of gridded biodiversity data.

```{r map-aoi}
ggplot() +
  geom_sf(data = aoi_grid, fill = NA, color = "darkgrey", alpha = 0.5) +
  geom_point(data = grid_spp,
             aes(x = centroid_lon, y = centroid_lat,
                 size = sqrt(obs_sum),
                 color = sqrt(obs_sum))) +
  scale_color_viridis_c(option = "turbo") +
  geom_sf(data = rsa, fill = NA, color = "black") +
  theme_minimal() +
  labs(title = "0.5° Grid with Observation Counts",
       x = "Longitude", y = "Latitude")
```

------------------------------------------------------------------------

### Example 1 - Species Richness

Here we calculate species richness across sites in the block_sp dataset, again using the compute_orderwise function. The richness function is applied to the grid_id column for site identification, with species data specified by sp_cols. Orders 1 to 4 are computed i.e. for order=1, it computes basic species richness at individual sites, while higher orders (2 to 4) represent the differences in richness between pairwise and/or multi-site combinations. A subset of 1000 samples is used for higher-order computations to speed-up computation time. Parallel processing is enabled with 4 worker threads to improve performance. The output is a table summarizing species richness across specified orders.

```{r compute-rich}
# Compute species richness (order 1) and the difference thereof for orders 2 to 4
rich_o1234 = compute_orderwise(
  df = grid_spp,
  func = richness,
  site_col = 'grid_id',
  sp_cols = sp_cols,
  sample_no = 1000,
  order = 1:4,
  parallel = TRUE,
  n_workers = 4)

head(rich_o1234)
```

```{r richness-plot}
# Plot species richness distribution by order
head(rich_o1234)
boxplot(value ~ order,
        data = rich_o1234,
        col = c('#4575b4', '#99ce8f', '#fefab8', '#d73027'),
        horizontal = TRUE,
        outline = FALSE,
        main = 'Distribution of Species Richness by Order')
rich_o1234 = as.data.frame(rich_o1234)
rich_o1234$centroid_lon = grid_spp$centroid_lon[match(rich_o1234$site_from, grid_spp$grid_id)]
rich_o1234$centroid_lat = grid_spp$centroid_lat[match(rich_o1234$site_from, grid_spp$grid_id)]

# Summarise turnover by site (spatial location)
mean_rich_o1234 = rich_o1234 %>%
  group_by(order, site_from, centroid_lon, centroid_lat) %>%
  summarize(value = mean(value, na.rm = TRUE))
head(mean_rich_o1234)
```
------------------------------------------------------------------------

### Example 2 - Community Turnover

Here we calculate species turnover (beta diversity) across sites in the block_sp dataset using the compute_orderwise function again. The turnover function is applied to the grid_id column for site identification, with species data specified by sp_cols. Order = 1 is not an option because turnover requires a comparison between sites. For orders 2 to 5, it computes turnover for pairwise and higher-order site combinations, representing the proportion of species not shared between sites. A subset of 1000 samples is used for higher-order comparisons. Parallel processing with 4 worker threads improves efficiency, and the output is a table summarizing species turnover across the specified orders.

```{r compute-turn}
# Compute community turnover for orders 2 to 5
turn_o2345 = compute_orderwise(
  df = grid_spp,
  func = turnover,
  site_col = 'grid_id',
  sp_cols = sp_cols, # OR `names(grid_spp)[-c(1:4)]`
  sample_no = 1000, # Reduce to speed-up computation
  order = 2:5,
  parallel = TRUE,
  n_workers = 4)

# Check results
head(turn_o2345)
```

To visualize the spatial patterns of turnover across sites, geographic coordinates are added back to the results. This allows spatial exploration of turnover patterns across different orders, highlighting regions of high or low turnover and enabling comparisons across orders. These visualizations provide valuable insights into spatial biodiversity dynamics.
Below we assign the geographic coordinates (x and y) from the block_sp dataset to the turn_o2345 results. Using match, it aligns the coordinates to the site_from column in turn_o2345 based on the corresponding grid_id values in block_sp. This prepares the dataset for spatial plotting.

```{r turnover-plot}
# Add coordinates back to 'turn_o2345' for plotting
# turn_o2345$x = grid_spp$x[match(turn_o2345$site_from, grid_spp$grid_id)]
# turn_o2345$y = grid_spp$y[match(turn_o2345$site_from, block_sp$grid_id)]

turn_o2345$centroid_lon = grid_spp$centroid_lon[match(turn_o2345$site_from, grid_spp$grid_id)]
turn_o2345$centroid_lat = grid_spp$centroid_lat[match(turn_o2345$site_from, grid_spp$grid_id)]

# Summarise turnover by site (spatial location)
mean_turn_o2345 = turn_o2345 %>%
  group_by(order, site_from, centroid_lon, centroid_lat) %>%
  summarize(value = mean(value, na.rm = TRUE))

# Plot Beta Diversity (pairwise turnover) from code above
ggplot() +
  geom_tile(data = mean_turn_o2345[mean_turn_o2345$order==2,],
            aes(x = centroid_lon, y = centroid_lat, fill = value)) +
  scale_fill_gradientn(colors = viridis(8)) + #Apply viridis color palette
  geom_sf(data = rsa, fill = NA, color = "black", alpha = 0.5) +
  theme_minimal() +
  labs(x = "Longitude", y = "Latitude", fill = "Beta Diversity") +
  theme(panel.grid = element_blank(),panel.border = element_blank()
  )

# Plot results to help visualise spatial patterns of turnover across different orders. This can highlight regions of high or low turnover and facilitate comparison across orders, providing insights into spatial biodiversity dynamics.
ggplot() +
  geom_tile(data = mean_turn_o2345, aes(x = centroid_lon, y = centroid_lat, fill = value)) +
  scale_fill_viridis_c(option = "turbo", name = "Turnover") +
  geom_sf(data = rsa, fill = NA, color = "black", alpha = 0.5) +
  theme_minimal() +
  labs(
    title = "Mean Turnover by Order",
    x = "Longitude",
    y = "Latitude"
  ) +
  facet_wrap(~ order, ncol = 2)
```

------------------------------------------------------------------------

### 7. Generate site by species matrix - `site_spp`

Create a matrix of species counts per site for use in biodiversity and dissimilarity analyses.

```{r site-spp}
xy = grid_spp[,2:3]
site_spp = grid_spp[,c(1:3,5:ncol(grid_spp))]
head(site_spp[,1:10])
dim(site_spp)
```

#### Generate record of occurrence counts called `sam_eff`

Summarize the number of species observations per grid cell to assess sampling effort.

```{r sam-eff}
sam_eff = grid_spp[, c("grid_id","centroid_lon","centroid_lat","obs_sum")]
obs_cnt = grid_spp[,c(1:3,5)]
dim(obs_cnt)
```

#### Generate a binary data frame 'sbs'

Convert species abundance data to presence--absence format (`site_spp_pa`) for binary dissimilarity analyses.

```{r sbs}
sp_cols = names(site_spp)[-(1:5)]
sbs     = site_spp %>%
  mutate(across(all_of(sp_cols), ~ ifelse(. > 0, 1, 0)))

# Assuming 'sp_cols' is a vector of column names
site_spp_pa = site_spp %>%
  mutate(across(all_of(sp_cols), ~ ifelse(!is.na(.) & . > 0, 1, 0)))

head(site_spp_pa[,1:10])
dim(site_spp_pa)
```

#### Generate richness of sbs row sums and map

Calculate species richness per site based on binary presence--absence data.

```{r spp-rich}
spp_rich = grid_spp[, c("grid_id","centroid_lon","centroid_lat","spp_rich")]
```

#### Generate 'sam.eff' and 'spp_rich' raster map

Visualise sampling effort and species richness as raster layers over the study area.

```{r eff-rich}
ras_effRich = grid_list$grid[[2:3]]
plot(sqrt(ras_effRich), col = turbo(100))
```

***Note**: occurrence coordinates are only used for assigning then into grids; they are not needed beyond this step.*

------------------------------------------------------------------------

### 8. Generate site by environment matrix using `get_enviro_data`

Use `get_enviro_data` to extract environmental variables for spatial points from either species observations or grid centroids.
Data can be sourced from online repositories (e.g. *WorldClim*, *SoilGrids* via `geodata`) or local rasters.
The function defines an area of interest with a buffer, extracts selected variables, and interpolates missing values using nearby non-NA points.
Outputs include cropped rasters, spatial points (as an `sf` object), and a combined site-by-environment data frame (`site_env`) for downstream ecological analyses.
Future support will extend to *CHELSA* (`climenv`), *Google Earth Engine* (`rgee`), and `mapme.biodiversity` for expanded environmental and biodiversity data access.

```{r get-enviro}
enviro_list = get_enviro_data(
  data      = sbs,
  buffer_km = 10,
  source    = 'geodata',
  var       = "bio",
  res       = 5,
  path      = 'download_data',
  sp_cols   = 6:ncol(sbs),
  ext_cols  = c('obs_sum','spp_rich')
)
ras_enviro = enviro_list$env_rast

# Optional: Rename the columns to something more descriptive
names(ras_enviro) = c("temp_mean", "mdr", "iso", "temp_sea", "temp_max",
                    "temp_min", "temp_rang","temp_wetQ","temp_dryQ", "temp_warmQ",
                    "temp_coldQ", "rain_mean","rain_wet", "rain_dry", "rain_sea",
                    "rain_wetQ", "rain_dryQ","rain_warmQ", "rain_coldQ")
```

#### Add 'eff-rich' raster to the enviro stack raster

Combine environmental rasters with sampling effort and species richness layers for joint spatial analysis.

```{r ras}
ras_enviro_effRich = c(ras_effRich, resample(ras_enviro, ras_effRich))
plot(ras_enviro_effRich[[1:4]])
```

#### Extract environmental variables for site centroids 'xy'

Link environmental data to grid centroids and visualize spatial variation in selected climate variables.

```{r env-df}
env_df = enviro_list$env_df[,c(1:5,8:26)]
head(env_df)
head(env_df[,15:24])

# Rename `env_df` column names
names(env_df) = c('grid_id','centroid_lon','centroid_lat','obs_sum','spp_rich',"temp_mean", "mdr", "iso", "temp_sea", "temp_max",
                                "temp_min", "temp_rang","temp_wetQ", "temp_dryQ", "temp_warmQ",
                                "temp_coldQ", "rain_mean","rain_wet", "rain_dry", "rain_sea",
                                "rain_wetQ", "rain_dryQ","rain_warmQ", "rain_coldQ")

# Check the first few rows
head(env_df)

# Plot the results to check the conversion is accurate
ggplot() +
  # Add 0.25deg grid layer
  geom_sf(data = aoi_grid, fill = NA, color = "darkgrey", alpha = 0.5) +
  # Add bfly points layer
  geom_point(data = env_df, aes(x = centroid_lon,
                                  y = centroid_lat,
                                  color = temp_mean)) +
  scale_size_continuous(range = c(0.5, 5)) +  # Adjust the point size range (min, max)
  scale_color_viridis_c(option = "turbo") +  # Use turbo color scale
  # Add boundary layer
  geom_sf(data = rsa, fill = NA, color = "black", alpha = 1) +
  theme_minimal() +
  labs(
    title = "0.25° Grid Cells with temp_mean",
    x = "Longitude",
    y = "Latitude")   
```

#### Generate *Site-by-Environment* data frame 'sbe'

Create a unified data frame of site coordinates, sampling effort, richness, and environmental variables for modelling called `sbe`.

```{r sbe}
sbe = env_df %>%
  select(grid_id, centroid_lon, centroid_lat, obs_sum, spp_rich, everything())
```

### 9. Change coordinates projection using `sf::st_transform`

Reproject spatial coordinates from geographic (WGS84) to a projected system (e.g. Albers Equal Area) for analyses requiring distance in meters, such as spatial clustering or environmental modelling.

```{r reproject}
# Convert to sf object with WGS84 geographic CRS
head(xy)
xy_sf = st_as_sf(xy, coords = c("centroid_lon", "centroid_lat"), crs = 4326)

# Project to Albers Equal Area (meters)
xy_utm = st_transform(xy_sf, crs = 9822)

# Extract transformed coordinates in meters
# Combine transformed coordinates back into data frame
xy_utm_df = cbind(xy, st_coordinates(xy_utm))
head(xy_utm_df)
```

------------------------------------------------------------------------

### 10. Check for colinearity using `rm_correlated`

Identify and remove highly correlated environmental variables to reduce multicollinearity in subsequent analyses.
This step ensures model stability by retaining only informative, non-redundant predictors based on a user-defined correlation threshold.

```{r remove_var}
# Remove the highly correlated variables
env_vars_reduced = rm_correlated(data = env_df[,c(4,6:24)],
                                 cols = NULL,
                                 threshold = 0.7,
                                 plot = TRUE)
head(env_vars_reduced)
dim(env_vars_reduced)
env_vars_reduced$obs_max = max(env_vars_reduced$obs_sum)
```

------------------------------------------------------------------------

### 11. Calculate Zeta decline for orders 2:15

#### Expectation of zeta diversity decline using `zetadiv::Zeta.decline.ex`

Computes the expectation of zeta diversity, the number of species shared by multiple assemblages for a range of orders (number of assemblages or sites), using a formula based on the occupancy of the species, and fits the decline to an exponential and a power law relationship.
*Generate statistics and figures, no maps*

```{r zeta-decline-ex}
zeta_decline_ex = Zeta.decline.ex(site_spp_pa[,6:ncol(site_spp_pa)], 
                                  orders = 1:15)
# zeta_decline_ex
```

> -   **Panel 1 (Zeta diversity decline)**: Shows how rapidly species that are common across multiple sites decline as you look at groups of more and more sites simultaneously (increasing zeta order). The sharp drop means fewer species are shared among many sites compared to just a few.
> -   **Panel 2 (Ratio of zeta diversity decline)**: Illustrates the proportion of shared species that remain as the number of sites compared increases. A steeper curve indicates that common species quickly become rare across multiple sites.
> -   **Panel 3 (Exponential regression)**: Tests if the decline in shared species fits an exponential decrease. A straight line here indicates that species commonness decreases rapidly and consistently as more sites are considered together. Exponential regression represents [**stochastic assembly**]{.underline} (**randomness determining species distributions**).
> -   **Panel 4 (Power law regression)**: Tests if the decline follows a power law relationship. A straight line suggests that the loss of common species follows a predictable pattern, where initially many species are shared among fewer sites, but rapidly fewer are shared among larger groups. Power law regression represents [**niche-based sorting**]{.underline} (**environmental factors shaping species distributions**).

> **Interpretation**: The near‐perfect straight line in the exponential panel (high R²) indicates that an exponential model provides the most parsimonious description of how species shared across sites decline as you add more sites—consistent with a stochastic, memory-less decline in common species. A power law will also fit in broad strokes, but deviates at high orders, suggesting exponential decay is the better choice for these data.

#### Zeta diversity decline using Monte Carlo sampling `zetadiv::Zeta.decline.mc`

Computes zeta diversity, the number of species shared by multiple assemblages, for a range of orders (number of assemblages or sites), using combinations of sampled sites, and fits the decline to an exponential and a power law relationship.
*Generate statistics and figures, no maps*

```{r zeta-decline-mc}
zeta_mc_utm = Zeta.decline.mc(site_spp_pa[,-(1:6)],
                               xy_utm_df[,3:4],
                               orders = 1:15,
                               sam = 100,
                               NON = TRUE,
                               normalize = "Jaccard")
# zeta_mc_utm
```

> -   **Panel 1 (Zeta diversity decline)**: Rapidly declining zeta diversity, similar to previous plots, indicates very few species remain shared across increasingly larger sets of sites, emphasizing strong species turnover and spatial specialization.
> -   **Panel 2 (Ratio of zeta diversity decline)**: More irregular fluctuations suggest a spatial effect: nearby sites might occasionally share more species by chance due to proximity. The spikes mean certain groups of neighboring sites have higher-than-average species overlap.
> -   **Panel 3 & 4 (Exponential and Power law regressions)**: Both remain linear, clearly indicating the zeta diversity declines consistently following a predictable spatial pattern. However, the exact pattern remains similar to previous cases, highlighting that despite spatial constraints, common species become rare quickly as more sites are considered.\
>     *This result demonstrates clear spatial structuring of biodiversity---species are locally clustered, not randomly distributed across the landscape. Spatial proximity influences which species co-occur more frequently. In practice use `Zeta.decline.mc` for real‐world biodiversity data—both because it scales and because the Monte Carlo envelope is invaluable when ζₖ gets noisier at higher orders.*

### 12. Calculate Zeta decay for orders 2:8

#### Zeta distance decay for a range of numbers of assemblages or sites using `zetadiv::Zeta.ddecays`

Computes the distance decay of zeta diversity for a range of orders (number of assemblages or sites), using generalised linear models.
*Generate statistics and figures, no maps*

```{r zeta-decays}
# zeta_decay = Zeta.ddecays(
#   xy      = xy[,c("centroid_lon","centroid_lat")],
#   data    = sbs[,-(1:5)],
#   sam     = 1000,
#   orders  = 2:15,
#   plot    = TRUE
# )

zeta_decays = Zeta.ddecays(xy_utm_df[,3:4],
                          site_spp_pa[,-(1:6)],
                          sam = 1000,
                          orders = 2:8,
                          plot = TRUE,
                          confint.level = 0.95
)
# zeta_decays
```

> This plot shows how zeta diversity (a metric that captures shared species composition among multiple sites) changes with spatial distance across different orders of zeta (i.e., the number of sites considered at once).

> -   On the **x-axis**, we have the **order of zeta** (from 2 to 7).\
>     For example, zeta order 2 looks at pairs of sites, order 3 at triplets, etc.
> -   On the **y-axis**, we see the slope of the **relationship between zeta diversity and distance** (i.e., how quickly species similarity declines with distance).
> -   A **negative slope** means that **sites farther apart have fewer species in common** --- so there's a clear distance decay of biodiversity.
> -   A **slope near zero** means **distance doesn't strongly affect how many species are shared among sites**.

> **Key observations:** - At low orders (2 and 3), the slope is strongly negative, indicating that species turnover is high over distance when looking at pairs or triplets of sites.
> - From order 4 and up, the slope becomes close to zero, suggesting that at broader spatial scales (more sites), species similarity is less affected by distance.\
> This may reflect widespread or core species that are consistently shared regardless of location.
> - The confidence intervals (error bars) shrink with increasing order, indicating greater stability and reliability of the estimate as more sites are included.

> **Summary:** This figure shows that biodiversity patterns across space are strongly shaped by distance at small scales, but this effect weakens as you include more sites.
> In other words, rare or localized species contribute to strong distance decay, but widespread species dominate at higher spatial scales, leading to more uniformity.
> *Species that occur in just two or three sites show a clear “farther apart → fewer shared species” pattern. But when you ask about the handful of really widespread species (those present in four, five, or more sites), their shared‐species counts no longer decline with distance—they form a spatially uniform core.*

### 13. Run a Multi-Site Generalised Dissimilarity Model for order 2

#### Multi-site generalised dissimilarity modelling for a set of environmental variables and distances using `zetadiv::Zeta.msgdm`

Computes a regression model of zeta diversity for a given order (number of assemblages or sites) against a set of environmental variables and distances between sites.
The different regression models available are generalised linear models, generalised linear models with negative constraints, generalised additive models, shape constrained additive models, and I-splines.
*Generate statistics and figures, no maps and save fitted order 2 model 'zeta2'*

```{r zeta-msgdm}
# zeta2 = Zeta.msgdm(
#   sbs[,-(1:5)],
#   sbe[,6:ncol(sbe)],
#   xy[,c("centroid_lon","centroid_lat")],
#   sam        = 1000,
#   order      = c(2,3,5,10),
#   distance.type = "Euclidean",
#   normalize  = "Jaccard",
#   r

# Compute a regression model of zeta diversity order 2
zeta2 = Zeta.msgdm(site_spp_pa[,-(1:6)],
                   env_vars_reduced[,-8],
                   # xy_utm_df[,3:4],
                   xy[,1:2],
                   sam = 1000,
                   order = 2,
                   distance.type = "Euclidean",# "ortho",
                   normalize = "Jaccard",
                   reg.type = "ispline")
# zeta2

# Compute splines coordinates from I-spline-based MSGDM
zeta2.ispline = Return.ispline(zeta2,
                               env_vars_reduced[,-8],
                               # legend = FALSE,
                               distance = TRUE)
# zeta2.ispline
Plot.ispline(isplines = zeta2.ispline, distance = TRUE)
```

> This figure shows the fitted I-splines from a multi-site generalized dissimilarity model (via `Zeta.msgdm`), which represent the partial, monotonic relationship between each predictor and community turnover (ζ-diversity) over its 0–1 “rescaled” range.  A few key take-aways:

> 1. **Distance (blue asterisks)** has by far the largest I-spline amplitude—rising from \~0 at zero distance to \~0.05 at the maximum.  That tells us spatial separation is the strongest driver of multi‐site turnover, and even small increases in distance yield a substantial drop in shared species.

> 2. **Sampling intensity (`obs_sum`, open circles)** comes next, with a gentle but steady rise to \~0.045.  This indicates that sites with more observations tend to share more species (or, conversely, that incomplete sampling can depress apparent turnover).

> 3. **Precipitation variables**: **Rain in the warm quarter (`rain_warmQ`, squares)** and **Rain in the dry quarter (`rain_dry`, triangles-down)** both show moderate effects (I-spline heights \~0.02–0.03).  This means differences in seasonal rainfall regimes contribute noticeably to changes in community composition.

> 4. **Temperature metrics**: **Mean temperature** *(`temp_mean`, triangles-up)*, **Wet‐quarter temperature** *(`temp_wetQ`, X’s)*, **Dry‐quarter temperature** *(`temp_dryQ`, diamonds)*, and the **isothermality index** *(`iso`, plus signs)* all have very low, almost flat I-splines (max heights ≲0.01).  In other words, these thermal variables explain very little additional turnover once you’ve accounted for distance and rainfall.

> **Ecological interpretation:** Spatial distance is the dominant structuring factor in these data—sites further apart share markedly fewer species.  After accounting for that, differences in observation effort and, to a lesser degree, seasonal rainfall still shape multisite community similarity.  Temperature and seasonality metrics, by contrast, appear to have only a minor independent influence on zeta‐diversity in this landscape.


```{r model-results}
# Deviance explained summary results
with(summary(zeta2$model), 1 - deviance/null.deviance) 
# [1] 0.2461561
# 0.2461561 means that approximately 24.6% of the variability in the response
# variable is explained by your model. This is relatively low, suggesting that the
# model may not be capturing much of the underlying pattern in the data.

# Model summary results
summary(zeta2$model)
```

------------------------------------------------------------------------

### 14. Predict current Zeta Diversity (zeta2) with `obs_sum` and current environmental variables

-   In the 'sbe' add 'sam.max', a constant for all sites = max(sam.eff)\
-   Predict for the updated sbe and xy\
-   Produce a site by site matrix of predicted zeta 'zeta.now' :::

```{r predict-zeta2}
# sbe_now = sbe %>% mutate(sam_max = max(obs_sum))
# zeta_now = Predict.zeta(zeta2, sbe_now, xy[,3:4])

# env_vars_reduced$obs_max = max(env_vars_reduced$obs_sum)
head(mean_rich_o1234)
head(mean_turn_o2345)

predictors_df = predict_dissim(
  block_sp   = grid_spp[,-c(4:6)],
  sbe        = sbe[,-c(1:3)],# env_vars_reduced[,-8]
  zeta_model = zeta2,
  mean_rich  = mean_rich_o1234,
  mean_turn  = mean_turn_o2345,
  sbs_xy     = xy,
  x_col      = "centroid_lon",
  y_col      = "centroid_lat",
  rsa        = rsa
)

head(predictors_df)
```

-   Run nmds for the predicted zeta matrix\
-   Plot RGB of the 3 component scores :::

------------------------------------------------------------------------

### 15. Run clustering analyses to map bioregion clusters of current zeta2

-   Generate maps of dissimilarity (the rgb plot)\
-   Generate map of bioregions (from clustering) :::

```{r zeta-cluster}
# e.g. perform clustering on zeta_now & plot bioregions
# Run `map_bioreg` function
# ‘data’: Input data containing coordinates (x, y) and environmental variables.
# ‘scale_cols’: Columns to be scaled and used for clustering.
# ‘clus_method’: Clustering method to apply ("kmeans", "pam", "hclust", "gmm", or "all" for all methods).
# ‘plot’: Logical flag; if TRUE, plots spatial maps of clustering results.
# ‘interp’: Specifies interpolation method ("NN", "Tps", "both") for filling spatial gaps using nearest neighbour (NN) or thin-plate splines (TPS).
head(predictors_df)
bioreg_result = map_bioreg(
  data = predictors_df,
  scale_cols = c("pred_zeta", "centroid_lon", "centroid_lat"),
  clus_method = "all", # K-means, PAM, Hierarchical and GMM clustering
  plot = TRUE,
  interp = "both",
  x_col ='centroid_lon',
  y_col ='centroid_lat'
)
```

------------------------------------------------------------------------

### 16. Predict future Zeta Diversity and map bioregion change

-   with appended (future scenarios) environmental variables and 'sam.max' in sbe\
-   For m number of scenarios plus the present scenario (step 14) and n sites of xy\
-   Updated sbe.future will have k = (m+1) x n number of rows\
-   'xy' also updated with k rows\
-   Predict a k by k matrix of predicted zeta 'zeta.future' :::

```{r predict-future}
# future_df = predict_dissim(
#   block_sp   = grid_spp[,-c(4:6)],
#   sbe        = sbe[,-c(1:3)],
#   zeta_model = zeta2,
#   mean_rich  = mean_rich_o1234,
#   mean_turn  = mean_turn_o2345,
#   sbs_xy     = sbs_xy,
#   x_col      = "centroid_lon",
#   y_col      = "centroid_lat",
#   rsa        = rsa
# )
# 
# head(future_df)

# Run `map_bioregDiff`
# Assuming 'bioreg_result$clusters' is a SpatRaster
# ‘raster_input’: A SpatRaster object with multiple layers to compare.
# approach: Specifies which metric to compute:
# >> "difference_count": Counts cells with differing values across layers
# >> "shannon_entropy": Calculates Shannon entropy, measuring diversity within layers
# >> "stability": Identifies stable regions where values remain unchanged.
# >> "transition_frequency": Quantifies the frequency of changes between layers
# >> "weighted_change_index": Weighted score to changes based on a dissimilarity matrix
# >> "all" (default): Returns all metrics as a combined SpatRaster
result_bioregDiff = map_bioregDiff(bioreg_result$clusters, approach = "all")
result_bioregDiff

# Mask the raster grid using the boundary of South Africa
# Convert to a terra vector
rsa_vect = vect(rsa)
# Create an empty raster over RSA at your desired resolution
# Here we use the extent of rsa_vect and a 0.05-degree cell size:
grid = rast(rsa_vect, resolution = 0.2489884, crs = crs(rsa_vect))
values(grid) = 1   # fill with dummy values

# Mask that grid to the RSA boundary
grid_masked = mask(grid, rsa_vect)
mask_bioregDiff = mask(resample(result_bioregDiff, grid_masked, method = "near"), grid_masked)

# Plot all change results
plot(mask_bioregDiff, col = viridis(100, direction = -1))
```

-   Nmds of zeta.future\
-   clustering of zeta.future\
-   Map sub matrices to indicate predicted future dissimilarity\
-   Map predicted future bioregions\
-   Map temporal turnover.\
    *Note: step 14 is redundant if step 16 is needed*\
    *Note: step 16 has the same code but more results including those from step 6 but potentially computational demanding*

------------------------------------------------------------------------

### 17. Deposit all results into Zenodo

All data frames, tables, maps, and standard metadata can be deposited into Zenodo using [`zen4R`](https://github.com/eblondel/zen4R/wiki) or [Zenodo UI](https://zenodo.org/records/10715460).

```{r zenodo}
# save(occ, sbs, sbe, zeta2, zeta_now, zeta_future, file="dissmapr_results.Rdata")
# use zen4R or Zenodo UI to upload archive
```

------------------------------------------------------------------------
