---
title: "dissmapr: Workflow for Compositional Dissimilarity & Biodiversity Turnover Analysis"
output:
  github_document:
    toc: true
    toc_depth: 2
  html_document:
    toc: true
    theme: cerulean     # or bootstrap, etc.
    mathjax: default
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%",
  echo = TRUE,
  message = FALSE,
  warning = FALSE
)
```

# `dissmapr`

<!-- badges: start -->
[![Binder](https://mybinder.org/badge_logo.svg)](https://mybinder.org/v2/gh/nithecs-biomath/RBasicPack/master?urlpath=rstudio)
[![Lifecycle: stable](https://img.shields.io/badge/lifecycle-stable-brightgreen.svg)](https://lifecycle.r-lib.org/articles/stages.html#stable)
[![test-coverage](https://github.com/macSands/dissmapr/actions/workflows/test-coverage.yaml/badge.svg)](https://github.com/macSands/dissmapr/actions/workflows/test-coverage.yaml)
[![Codecov test coverage](https://codecov.io/gh/macSands/dissmapr/graph/badge.svg)](https://app.codecov.io/gh/macSands/dissmapr)
[![R-CMD-check](https://github.com/macSands/dissmapr/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/macSands/dissmapr/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

# `dissmapr`: A Novel Framework for Automated Compositional Dissimilarity and Biodiversity Turnover Analysis

## Introduction

**dissmapr** provides a reproducible, end-to-end workflow for computing and mapping compositional dissimilarity and biodiversity turnover across large spatial and temporal scales. Core modules handle occurrence retrieval, grid generation, environmental extraction, order-wise dissimilarity, and bioregional mapping.
---

## Tutorial

This document outlines a step-by-step workflow for analyzing compositional dissimilarity and bioregionalization using **dissmapr**.  
All code chunks below illustrate how one might implement each step in R, using various commonly used packages (e.g., `sf`, `terra`, `dplyr`, `ggplot2`, etc.).  
Please note that paths to data, package names, or exact functions may need adjustment depending on your local setup.
---

## Installation

```{r install, eval=FALSE, include=FALSE}
# install remotes if needed
# install.packages("remotes")
remotes::install_github("macSands/dissmapr")
```
---

## Setup

```{r load-dissmapr}
# Load your package 
# library(dissmapr)
library(dissmapr)

# Make sure so all exported functions are available
devtools::load_all()

# setwd('D:\\Methods\\R\\myR_Packages\\myCompletePks\\dissmapr')
```
---

## Load libraries

```{r libraries}
# Load necessary libraries
library(httr)       # HTTP client  
library(geodata)    # Download geographic data  
library(data.table) # Fast large-table operations  
library(dplyr)      # Data manipulation verbs  
library(tidyr)      # Tidy data reshaping  
library(zoo)        # Time series utilities  
library(sf)         # Vector spatial data  
library(terra)      # Raster spatial operations  
library(zetadiv)    # Multi-site biodiversity turnover  
library(ggplot2)    # Grammar of graphics  
library(viridis)    # Perceptual color scales  
```
---

## 1. User-defined area of interest and grid resolution 
### Read RSA shape file 
```{r aoi}
rsa = sf::st_read('inst/extdata/rsa.shp')
```
---

## 2. Site by species matrix and sampling effort 
### Access occurrences of user-specified taxon or species list using `get_occurrence_data`
```{r get-occurrence}
bfly_data = get_occurrence_data(
  data        = 'inst/extdata/gbif_butterflies.csv',
  source_type = 'local_csv',
  sep         = '\t'
)
dim(bfly_data)
```
---

## 3. Format data using `format_df`
### Format data into long and wide formats
```{r format-df}
bfly_result = format_df(
  data        = bfly_data,
  species_col = 'verbatimScientificName',
  value_col   = 'pa',
  extra_cols  = NULL,
  format      = 'long'
)
site_obs = bfly_result$site_obs
site_spp = bfly_result$site_sp
dim(site_obs)
dim(site_spp)
```

### Get parameters to use later
```{r spp-cols}
# Get parameters from processed data
# Number of species columns
(n_sp = dim(site_spp)[2] - 3)
sp_cols = names(site_spp)[-c(1:3)]
```
---

## 4. Summarise records by grid using `generate_grid`
### Assign records to a grid at a set resolution (e.g. 0.5°)
```{r grid}
grid_list = generate_grid(
  data       = site_spp,
  x_col      = "x",
  y_col      = "y",
  grid_size  = 0.5,
  sum_col_range = 4:ncol(site_spp),
  crs_epsg   = 4326
)

aoi_grid = grid_list$grid_sf
grid_spp = grid_list$block_sp
dim(aoi_grid)
dim(grid_spp)
```

### Generate a data frame ‘xy’ of site centroids
```{r site-xy}
# Grid centroids with 'gird_id', 'centroid_lon', 'centroid_lat', 'obs_sum' and `spp_rich`
grid_xy = grid_spp[,c(1:3,5:6)]
dim(grid_xy)
spp_obs = site_obs
dim(spp_obs)
```

### Generate a map of RSA with occupied grid cells as centroid points
```{r map-aoi}
ggplot() +
  geom_sf(data = aoi_grid, fill = NA, color = "darkgrey", alpha = 0.5) +
  geom_point(data = grid_spp,
             aes(x = centroid_lon, y = centroid_lat,
                 size = sqrt(obs_sum),
                 color = sqrt(obs_sum))) +
  scale_color_viridis_c(option = "turbo") +
  geom_sf(data = rsa, fill = NA, color = "black") +
  theme_minimal() +
  labs(title = "0.5° Grid with Observation Counts",
       x = "Longitude", y = "Latitude")
```
---

## 5. Summarise records by grid using `generate_grid`
#### Generate site by species matrix \>\> `site_spp`
```{r site-spp}
xy = grid_spp[,2:3]
site_spp = grid_spp[,c(1:3,5:ncol(grid_spp))]
head(site_spp[,1:10])
dim(site_spp)
```

### Generate a raster of occurrence counts called `sam_eff`
```{r sam-eff}
sam_eff = grid_spp[, c("grid_id","centroid_lon","centroid_lat","obs_sum")]
obs_cnt = grid_spp[,c(1:3,5)]
dim(obs_cnt)
```

### Generate a binary data frame ‘sbs’
```{r sbs}
sp_cols = names(site_spp)[-(1:5)]
sbs     = site_spp %>%
  mutate(across(all_of(sp_cols), ~ ifelse(. > 0, 1, 0)))

# Assuming 'sp_cols' is a vector of column names
site_spp_pa = site_spp %>%
  mutate(across(all_of(sp_cols), ~ ifelse(!is.na(.) & . > 0, 1, 0)))

head(site_spp_pa[,1:10])
dim(site_spp_pa)
```

### Generate richness of sbs row sums and map
```{r spp-rich}
spp_rich = grid_spp[, c("grid_id","centroid_lon","centroid_lat","spp_rich")]
```

### Generate ‘sam.eff’ and ‘spp_rich’ raster map
```{r eff-rich}
ras_effRich = grid_list$grid[[2:3]]
plot(sqrt(ras_effRich), col = turbo(100))
```

### Note: occurrence coordinates are only used for assigning then into grids; they are not needed beyond this step.

---

## 6. Site by environment matrix
### Stack environmental variables into raster using `get_enviro_data`
```{r get-enviro}
enviro_list = get_enviro_data(
  data      = sbs,
  buffer_km = 10,
  source    = 'geodata',
  var       = "bio",
  res       = 5,
  path      = 'download_data',
  sp_cols   = 6:ncol(sbs),
  ext_cols  = c('obs_sum','spp_rich')
)
# ras_enviro = enviro_list$env_rast

# Rename the layers
names(enviro_list$env_rast) = c("temp_mean", "mdr", "iso", "temp_sea", "temp_max",
                      "temp_min", "temp_rang","temp_wetQ", "temp_dryQ", "temp_warmQ",
                      "temp_coldQ", "rain_mean","rain_wet", "rain_dry", "rain_sea",
                      "rain_wetQ", "rain_dryQ","rain_warmQ", "rain_coldQ")

ras_enviro = enviro_list$env_rast
```

### Add ‘eff-rich’ raster to the enviro stack raster
```{r ras}
ras_enviro_effRich = c(ras_effRich, resample(ras_enviro, ras_effRich))
plot(ras_enviro_effRich[[1:4]])
```

### Extract environmental variables for site centroids ‘xy’ 
```{r env-df}
env_df = enviro_list$env_df[,c(1:5,8:26)]
head(env_df)
head(env_df[,15:24])
names(env_df) = c('grid_id','x','y','obs_sum','spp_rich',"temp_mean", "mdr", "iso", "temp_sea", "temp_max",
                                "temp_min", "temp_rang","temp_wetQ", "temp_dryQ", "temp_warmQ",
                                "temp_coldQ", "rain_mean","rain_wet", "rain_dry", "rain_sea",
                                "rain_wetQ", "rain_dryQ","rain_warmQ", "rain_coldQ")

head(env_df)
ggplot() +
  # Add 0.25deg grid layer
  geom_sf(data = aoi_grid, fill = NA, color = "darkgrey", alpha = 0.5) +
  # Add bfly points layer
  geom_point(data = env_df, aes(x = x,
                                  y = y,
                                  color = temp_mean)) +
  scale_size_continuous(range = c(0.5, 5)) +  # Adjust the point size range (min, max)
  scale_color_viridis_c(option = "turbo") +  # Use turbo color scale
  # Add boundary layer
  geom_sf(data = rsa, fill = NA, color = "black", alpha = 1) +
  theme_minimal() +
  labs(
    title = "0.25° Grid Cells with temp_mean",
    x = "Longitude",
    y = "Latitude")   
```

### Generate data frame ‘sbe’
Extract environmental variables and sampling effort and species richness `ras_enviro` for site centroids `xy`
```{r sbe}
sbe = env_df %>%
  select(grid_id, x, y, obs_sum, spp_rich, everything())
```

## 7. If necessary project coordinates into meters projection (e.g. UTM)
```{r reproject}
# Convert to sf object with WGS84 geographic CRS
head(xy)
xy_sf = st_as_sf(xy, coords = c("centroid_lon", "centroid_lat"), crs = 4326)

# Project to Albers Equal Area (meters)
xy_utm = st_transform(xy_sf, crs = 9822)

# Extract transformed coordinates in meters
# Combine transformed coordinates back into data frame
xy_utm_df = cbind(xy, st_coordinates(xy_utm))
head(xy_utm_df)
```

## 8. Check for inter-correlation and remove highly corrected variables
```{r remove_var}
# Remove the highly correlated variables
env_vars_reduced = rm_correlated(data = env_df[,c(4,6:24)],
                                 cols = NULL,
                                 threshold = 0.7,
                                 plot = TRUE)
head(env_vars_reduced)
dim(env_vars_reduced)
```
---

## 9. Zeta decline (sbs), orders 2:15
### Expectation of zeta diversity decline using `zetadiv::Zeta.decline.ex`
#### Generate statistics and figures, no maps
```{r zeta-decline-ex}
zeta_decline_ex = Zeta.decline.ex(site_spp_pa[,6:ncol(site_spp_pa)], 
                                  orders = 1:15)
# zeta_decline_ex
```
> > -   **Panel 1 (Zeta diversity decline)**: Shows how rapidly species that are common across multiple sites decline as you look at groups of more and more sites simultaneously (increasing zeta order). The sharp drop means fewer species are shared among many sites compared to just a few.
> >
> > -   **Panel 2 (Ratio of zeta diversity decline)**: Illustrates the proportion of shared species that remain as the number of sites compared increases. A steeper curve indicates that common species quickly become rare across multiple sites.
> >
> > -   **Panel 3 (Exponential regression)**: Tests if the decline in shared species fits an exponential decrease. A straight line here indicates that species commonness decreases rapidly and consistently as more sites are considered together. Exponential regression represents [**stochastic assembly**]{.underline} (**randomness determining species distributions**).
> >
> > -   **Panel 4 (Power law regression)**: Tests if the decline follows a power law relationship. A straight line suggests that the loss of common species follows a predictable pattern, where initially many species are shared among fewer sites, but rapidly fewer are shared among larger groups. Power law regression represents [**niche-based sorting**]{.underline} (**environmental factors shaping species distributions**).

### Zeta diversity decline using Monte Carlo sampling `zetadiv::Zeta.decline.mc`
#### Generate statistics and figures, no maps
```{r zeta-decline-mc}
zeta_mc_utm = Zeta.decline.mc(site_spp_pa[,-(1:6)],
                               xy_utm_df[,3:4],
                               orders = 1:15,
                               sam = 100,
                               NON = TRUE,
                               normalize = "Jaccard")
# zeta_mc_utm
```
> > -   **Panel 1 (Zeta diversity decline)**: Rapidly declining zeta diversity, similar to previous plots, indicates very few species remain shared across increasingly larger sets of sites, emphasizing strong species turnover and spatial specialization.
> >
> > -   **Panel 2 (Ratio of zeta diversity decline)**: More irregular fluctuations suggest a spatial effect: nearby sites might occasionally share more species by chance due to proximity. The spikes mean certain groups of neighboring sites have higher-than-average species overlap.
> >
> > -   **Panel 3 & 4 (Exponential and Power law regressions)**: Both remain linear, clearly indicating the zeta diversity declines consistently following a predictable spatial pattern. However, the exact pattern remains similar to previous cases, highlighting that despite spatial constraints, common species become rare quickly as more sites are considered.
>
> **This result demonstrates clear spatial structuring of biodiversity—species are locally clustered, not randomly distributed across the landscape. Spatial proximity influences which species co-occur more frequently.**
---

## 10. Zeta decays (sbs, xy), orders 2:15
### Zeta distance decay for a range of numbers of assemblages or sites using `zetadiv::Zeta.ddecays`
#### Generate statistics and figures, no maps
```{r zeta-decays}
# zeta_decay = Zeta.ddecays(
#   xy      = xy[,c("centroid_lon","centroid_lat")],
#   data    = sbs[,-(1:5)],
#   sam     = 1000,
#   orders  = 2:15,
#   plot    = TRUE
# )

zeta_decays = Zeta.ddecays(xy_utm_df[,3:4],
                          site_spp_pa[,-(1:6)],
                          sam = 1000,
                          orders = 2:7,
                          plot = TRUE,
                          confint.level = 0.95
)
# zeta_decays
```
> > This plot shows how zeta diversity (a metric that captures shared species composition among multiple sites) changes with spatial distance across different orders of zeta (i.e., the number of sites considered at once).
> >
> > -   On the **x-axis**, we have the **order of zeta** (from 2 to 7). For example, zeta order 2 looks at pairs of sites, order 3 at triplets, etc.
> > -   On the **y-axis**, we see the slope of the **relationship between zeta diversity and distance** (i.e., how quickly species similarity declines with distance).
> > -   A **negative slope** means that **sites farther apart have fewer species in common**—so there's a clear distance decay of biodiversity.
> > -   A **slope near zero** means **distance doesn’t strongly affect how many species are shared among sites**.

> **Key observations:** - At low orders (2 and 3), the slope is strongly negative, indicating that species turnover is high over distance when looking at pairs or triplets of sites. - From order 4 and up, the slope becomes close to zero, suggesting that at broader spatial scales (more sites), species similarity is less affected by distance. This may reflect widespread or core species that are consistently shared regardless of location. - The confidence intervals (error bars) shrink with increasing order, indicating greater stability and reliability of the estimate as more sites are included.

> **Summary:** This figure shows that biodiversity patterns across space are strongly shaped by distance at small scales, but this effect weakens as you include more sites. In other words, rare or localized species contribute to strong distance decay, but widespread species dominate at higher spatial scales, leading to more uniformity. 
---

## 11. Zeta.msgdm(sbs, sbe, xy), order 2, 3, 5, 10
### Multi-site generalised dissimilarity modelling for a set of environmental variables and distances using `zetadiv::Zeta.msgdm`
#### Generate statistics and figures, no maps and save fitted order 2 model ‘zeta2’
```{r zeta-msgdm}
# zeta2 = Zeta.msgdm(
#   sbs[,-(1:5)],
#   sbe[,6:ncol(sbe)],
#   xy[,c("centroid_lon","centroid_lat")],
#   sam        = 1000,
#   order      = c(2,3,5,10),
#   distance.type = "Euclidean",
#   normalize  = "Jaccard",
#   r

# Compute a regression model of zeta diversity order 2
zeta2 = Zeta.msgdm(site_spp_pa[,-(1:6)],
                   env_vars_reduced,
                   xy_utm_df[,3:4],
                   sam = 1000,
                   order = 2,
                   distance.type = "Euclidean",# "ortho",
                   normalize = "Jaccard",
                   reg.type = "ispline")
# zeta2

# Compute splines coordinates from I-spline-based MSGDM
zeta2.ispline = Return.ispline(zeta2,
                               env_vars_reduced,
                               # legend = FALSE,
                               distance = TRUE)
# zeta2.ispline
Plot.ispline(isplines = zeta2.ispline, distance = TRUE)
```
>> This I-spline plot represents how different environmental variables (including distance) contribute to explaining zeta diversity of order 2 — that is, how many species are shared between pairs of sites, based on environmental similarity and geographic distance.

**Imagine you're comparing pairs of locations and asking: "Do these two places have similar species, and if so, why?"**
>> How to Read the Graph: This graph shows which factors matter most for explaining how similar the species are between those pairs of places.
\- X-axis (Rescaled range): Each variable has been normalized from 0 to 1, so we can compare their effects on the same scale.
\- Y-axis (I-splines): This shows how much each variable contributes to the similarity in species between two locations.
\- A steeper or higher curve = more important variable.

>> **Main Takeaways from the Plot**
-   Geographic **distance** (blue line) has the strongest effect:
-   As distance increases (from 0 to 1), species similarity drops — locations far apart share fewer species.
-   This confirms classic distance decay in ecology.
-   The **obs_sum** variable (likely representing total observed richness or sampling effort per site) shows a high and rapidly increasing I-spline curve, which then levels off but remains high across the rescaled range.
-   Strong initial effect: Differences in species richness between sites are a key driver of how many species they share — especially when the difference is large.
-   Plateau effect: After a certain point, increasing obs_sum further doesn't increase shared species much more, possibly due to saturation in species detection or common species already being captured.
-   Overall importance: Because the curve remains high, this variable is consistently influential across its range.
-   **Temperature** and **rainfall seasonality** also matter, but less: Variables like temp_mean, rain_warmQ, and temp_wetQ contribute, but their curves are lower. This suggests that climate seasonality explains some turnover in species, though not as strongly as distance.
-   Other climatic variables (e.g., isothermality, coldest quarter rain) have much smaller contributions. These environmental gradients do affect biodiversity, but only slightly.

In summary, besides geographic distance, the total number of species observed at each site (obs_sum) has a strong influence on how similar two places are in terms of species. When one site has many more species than another, they tend to share fewer species. However, once species richness becomes high enough, its additional effect on shared species plateaus.

```{r model-results}
# Deviance explained summary results
with(summary(zeta2$model), 1 - deviance/null.deviance) 
# [1] 0.04414301
# 0.04414301 means that approximately 4.41% of the variability in the response
# variable is explained by your model. This is relatively low, suggesting that the
# model may not be capturing much of the underlying pattern in the data.

# Model summary results
summary(zeta2$model)
```
---

## 12. Predict(zeta2) with ‘sam.eff’
### In the ‘sbe’ add ‘sam.max’, a constant for all sites = max(sam.eff)
### Predict for the updated sbe and xy 
### Produce a site by site matrix of predicted zeta ‘zeta.now’
```{r predict-zeta2}
# sbe_now = sbe %>% mutate(sam_max = max(obs_sum))
# zeta_now = Predict.zeta(zeta2, sbe_now, xy[,3:4])
```

### Run nmds for the predicted zeta matrix 
### Plot RGB of the 3 component scores

## 13. Clustering analyses directly using zeta.now
### Generate maps of dissimilarity (the rgb plot)
### Generate map of bioregions (from clustering)
```{r zeta-cluster}
# e.g. perform clustering on zeta_now & plot bioregions
```

---

## 14. Predict(zeta2) 
### with appended (future scenarios) environmental variables and ‘sam.max’ in sbe
### For m number of scenarios plus the present scenario (step 6a) and n sites of xy
### Updated sbe.future will have k = (m+1) x n number of rows
### ‘xy’ also updated with k rows
### Predict a k by k matrix of predicted zeta ‘zeta.future’
```{r predict-future}
# e.g. perform clustering on zeta_now & plot bioregions
```

### Nmds of zeta.future
### clustering of zeta.future
### Map sub matrices to indicate predicted future dissimilarity
### Map predicted future bioregions
### Map temporal turnover. 
### Note: step 6a is redundant if step 7 is needed
### Note: step 7 has the same code but more results including those from step 6 but potentially computational demanding

---

## 15. Deposit all data frames, tables, maps, and standard metadata to zenodo
```{r zenodo}
# save(occ, sbs, sbe, zeta2, zeta_now, zeta_future, file="dissmapr_results.Rdata")
# use zen4R or Zenodo UI to upload archive
```

---
