---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%",
  echo = TRUE,
  message = FALSE,
  warning = FALSE
)
```

<!-- badges: start -->

[![repo status](https://www.repostatus.org/badges/latest/active.svg)](https://www.repostatus.org/#active)
[![Release](https://img.shields.io/github/release/b-cubed-eu/dissmapr.svg)](https://github.com/b-cubed-eu/dissmapr/releases)
[![dissmapr status
badge](https://b-cubed-eu.r-universe.dev/dissmapr/badges/version)](https://b-cubed-eu.r-universe.dev/dissmapr)
[![CRAN status](https://www.r-pkg.org/badges/version/dissmapr)](https://cran.r-project.org//package=dissmapr)
[![R-CMD-check](https://github.com/b-cubed-eu/dissmapr/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/b-cubed-eu/dissmapr/actions/workflows/R-CMD-check.yaml)
[![Codecov test coverage](https://codecov.io/gh/b-cubed-eu/dissmapr/graph/badge.svg)](https://app.codecov.io/gh/b-cubed-eu/dissmapr)
[![DOI](https://img.shields.io/badge/DOI-awaiting_upload_to_zenodo-orange)](https://zenodo.org)
[![name status badge](https://b-cubed-eu.r-universe.dev/badges/:name?color=6CDDB4)](https://b-cubed-eu.r-universe.dev/)
[![MIT](https://img.shields.io/badge/License-MIT-yellow.svg)](LICENSE.md)

<!-- badges: end -->

# dissmapr

## A Novel Framework for Automated Compositional Dissimilarity and Biodiversity Turnover Analysis

## Introduction

Quantifying **dissimilarity** is central to ecology and biodiversity science. Whether comparing communities across landscapes, examining functional distances between species, or assessing environmental contrasts, dissimilarity metrics underpin analyses of diversity, turnover, and ecological change. However, these metrics are often scattered across packages, calculated inconsistently, or tailored to a single data type (e.g., species composition only).

**`dissmapr`** provides a unified, transparent, and reproducible workflow for calculating and analysing dissimilarities in biodiversity composition. The package supports **data acquisition**, **environmental linking**, **computation of order-wise dissimilarity**, and **mapping of bioregional patterns**, leveraging tools like `zetadiv` (for multi-site generalized dissimilarity modelling, MS-GDM), `terra`, `sf`, and `ggplot2`.

By combining flexible metric choice with consistent data handling and mapping functions, `dissmapr` is designed to streamline biodiversity workflows, support cross-study comparability, and provide decision-relevant outputs for conservation and global change research.

## Core Concepts

* **Occurrence data** – harmonising raw biodiversity records and linking them to spatial grids.
* **Environmental data** – retrieving, processing, and attaching environmental predictors to sites.
* **Order-wise dissimilarity** – extending beyond pairwise turnover to higher-order zeta diversity metrics.
* **Predictive modelling** – using MS-GDM to model compositional turnover across spatial and environmental gradients.
* **Bioregional mapping** – clustering and interpolating sites to delineate bioregions and track changes.

Together, these concepts provide a workflow for exploring biodiversity structure across ecological, functional, and environmental dimensions.

## Main Functions

The package is structured around modular steps, so users can build end-to-end workflows or extract individual components.

### **Data Handling**
* `get_occurrence_data()` – Import and harmonise biodiversity-occurrence data.  
* `generate_grid()` – Generate spatial grids and gridded summaries.  
* `assign_mapsheet()` – Add nearest mapsheet codes and centre coordinates.  
* `get_enviro_data()` – Retrieve, crop, resample, and link environmental rasters to sites.  
* `format_df()` – Format biodiversity records to long/wide forms for analysis.  

### **Dissimilarity & Metrics**
* `compute_orderwise()` – Compute order-wise metrics, including zeta diversity and dissimilarities.  
* `rm_correlated()` – Remove highly correlated predictors to avoid redundancy.  
* Helper metrics available via `helper_indices`, e.g.:  
  - `geodist_helper()` – Geographic distance (Haversine).  
  - `richness()` – Species richness.  
  - `turnover()` – Species turnover/beta diversity.  
  - `diss_bcurt()` – Bray–Curtis dissimilarity.  
  - `orderwise_diss_gower()` – Gower dissimilarity.  

### **Prediction & Mapping**
* `predict_dissim()` – Predict pairwise compositional turnover (zeta-dissimilarity) with richness.  
* `map_bioreg()` – Raster-based clustering and interpolation of bioregional data.  
* `map_bioregDiff()` – Map bioregional change metrics between categorical raster layers.  

### **ζ-MSGDM Workflow**
Functions to automate multi-site GDM workflows:
* `run_ispline_models()` – Fit multiple `zetadiv::Zeta.msgdm` I-spline models.  
* `plot_ispline_lines()` – Plot I-spline partial effects.  
* `plot_ispline_boxplots()` – Plot facetted boxplots for I-spline basis functions.  

## Applications

`dissmapr` is designed for use across a wide range of biodiversity research and conservation contexts, including:

* **Community ecology** – comparing species composition across landscapes, ecoregions, or time periods.  
* **Macroecology** – exploring turnover and nestedness at regional to global scales.  
* **Environmental filtering** – quantifying how abiotic differences shape community assembly.  
* **Conservation planning** – mapping bioregions and their shifts to identify hotspots, refugia, or vulnerable sites.  

## Why Use `dissmapr`?

* **Unified workflow** – integrates occurrence, environment, dissimilarity, and mapping in a consistent manner.  
* **Transparent & reproducible** – modular design supports repeatable workflows across datasets and studies.  
* **Interpretable outputs** – order-wise dissimilarities, predictive models, and mapped bioregions.  
* **Flexible & extensible** – built on standard R data structures and interoperable with packages like `zetadiv`, `terra`, `sf`, and `ggplot2`.  

## Installation

Install from GitHub using:

```{r}
# install.packages("remotes")
remotes::install_github("b-cubed-eu/dissmapr")
```

## Minimal Example

Below is a short workflow showing how to move from **occurrence data** to **order-wise dissimilarity** and **bioregional mapping**.

```{r}
library(dissmapr)

# 1. Import or simulate occurrence data (here using built-in example dataset)
occ = get_occurrence_data(example = TRUE)

# 2. Generate a grid for spatial aggregation
grid = generate_grid(occ, res = 0.5)

# 3. Format the data for analysis
df = format_df(occ, grid)

# 4. Compute order-wise dissimilarities (zeta diversity, turnover, etc.)
diss = compute_orderwise(df, func = richness)

# 5. Map bioregions using clustering and interpolation
bio_map = map_bioreg(diss, grid)

# 6. Plot the resulting bioregionalisation
plot(bio_map)
```

This workflow illustrates the package’s modular design: each function can be run independently or chained together for end-to-end biodiversity workflows.

## Citation

If you use **`dissmapr`**, please cite the package and associated methods. See `citation("dissmapr")` and the repository’s CITATION files.
