---
title: "1-setup"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{1-setup}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "figures/1-",
  out.width = "100%",
  echo = TRUE,
  message = FALSE,
  warning = FALSE
)
```

------------------------------------------------------------------------

# `dissmapr`

## A Novel Framework for Automated Compositional Dissimilarity and Biodiversity Turnover Analysis

------------------------------------------------------------------------

## Step-by-Step Workflow

This section provides a concise, reproducible “from-scratch” pathway for setting up **`dissmapr`**.  By the end of these steps you will have:

1. **Installed and loaded** the latest development version of `dissmapr` from GitHub.  
2. **Loaded all supporting packages** needed for spatial manipulation, biodiversity modelling, and visualisation.  
3. **Downloaded exemplar occurrence data** (South African butterflies) with `get_occurrence_data()`, demonstrating how to pull records from local files, remote URLs, or GBIF archives.  
4. **Tidied the records** into long and wide analytical formats via `format_df()`—crucially *without changing the native spatial resolution* of the observations.  
5. **Defined a study extent and analysis grid**, creating a masked raster template that will act as the spatial backbone for downstream compositional dissimilarity and turnover analyses.

Execute each chunk in the order presented (the chunk labels mirror the headings) or adapt them to your own taxonomic group, area of interest, or grid resolution.  All subsequent vignettes assume you have completed this workflow and have the resulting `site_obs`, `site_spp`, and grid objects in your workspace.

------------------------------------------------------------------------

### Install and load `dissmapr`

Install and load the `dissmapr` package from GitHub, ensuring all functions are available for use in the workflow.

```{r install, eval=FALSE, include=TRUE}
# install remotes if needed
# install.packages("remotes")
remotes::install_github("macSands/dissmapr")
```

```{r load-dissmapr, include=TRUE}
# Ensure the package is loaded when knitting
library(dissmapr)

# Make sure all the functions are loaded
devtools::load_all()
```

------------------------------------------------------------------------

### Load other R libraries

Load core libraries for spatial processing, biodiversity modelling, and visualization required across the `dissmapr` analysis pipeline.

```{r libraries}
# Load necessary libraries
library(httr)       # HTTP client  
library(geodata)    # Download geographic data  
library(data.table) # Fast large-table operations  
library(dplyr)      # Data manipulation verbs  
library(tidyr)      # Tidy data reshaping  
library(zoo)        # Time series utilities  
library(sf)         # Vector spatial data  
library(terra)      # Raster spatial operations  
library(tidyterra)  # supplies geom_spatraster()
library(zetadiv)    # Multi-site dissimilarity modelling
library(ggplot2)    # Grammar of graphics  
library(viridis)    # Perceptual color scales  
```

------------------------------------------------------------------------

### Get species occurrence records using `get_occurrence_data()`

To contextualise the following steps of the workflow, we use the butterfly data in South Africa, accessed from GBIF, as the case for demonstration. Ultimately, the choice for the AoI and taxa is user-specific. 
This section focuses on automating the retrieval and pre-processing of biodiversity occurrence data from various sources, including:

*  **local** `databases`.csv` files,
*  **URLs** or `.zip` files from the Global Biodiversity Information Facility (GBIF), and
*  species **occurrence cubes** from B3 (specification) [*work in progress*].

The function assembles data on species distributions across specified taxonomic groups and regions, producing presence-absence or abundance matrices that quantify species co-occurrence within locations.

```{r get-occurrence}
bfly_data = get_occurrence_data(
  data        = system.file("extdata", "gbif_butterflies.csv", package = "dissmapr"),
  source_type = 'local_csv',
  sep         = '\t'
)

# Check results
dim(bfly_data)
str(bfly_data)
head(bfly_data[,1:6])
```

------------------------------------------------------------------------

### Format data using `format_df()`

Use `format_df` to *standardise and reshape* raw biodiversity tables into the *long* or *wide* format required by later `dissmapr` steps.
Importantly, this function does not alter the spatial resolution of the original observations - it simply tidies the data by automatically identifying key columns (e.g., coordinates, species, and values), assigning missing observation/site IDs, and reformatting the data for analysis.
Outputs include a cleaned `site_obs` dataset and `site_spp` matrix for further processing:

*  **site_obs**: Simplified table with unique `site_id`, `x`, `y`, `species` and `value` records (long format).
*  **site_spp**: Site-by-species matrix for biodiversity assessments (wide format).

#### Format data into long (site_obs) and wide (site_spp) formats

```{r format-df}
bfly_result = format_df(
  data        = bfly_data, # A `data.frame` of biodiversity records
  species_col = 'verbatimScientificName', # Name of species column (required for `"long"`)
  value_col   = 'pa', # Name of value column (e.g. presence/abundance; for `"long"`)
  extra_cols  = NULL, # Character vector of other columns to keep
  format      = 'long' # Either`"long"` or `"wide"`. If `NULL`, inferred from `species_col` & `value_col`
)

# Check `bfly_result` structure
str(bfly_result, max.level = 1)

# Optional: Create new objects from list items
site_obs = bfly_result$site_obs
site_spp = bfly_result$site_spp

# Check results
dim(site_obs)
head(site_obs)

dim(site_spp)
head(site_spp[,1:6])

#### Get parameters from processed data to use later
# Number of species
(n_sp = dim(site_spp)[2] - 3)

# Species names
sp_cols = names(site_spp)[-c(1:3)]
sp_cols[1:10]
```
------------------------------------------------------------------------

### User-defined area of interest and grid resolution

Load the spatial boundary data for South Africa to serve as the geographic reference for all subsequent biodiversity analyses and visualizations.

```{r aoi}
# Read RSA shape file
rsa = sf::st_read(system.file("extdata", "rsa.shp", package = "dissmapr"))

# Define your resolution and create mask to use later
res = 0.5 # 0.5 degrees is roughly 55km

# Convert to a terra vector
rsa_vect = vect(rsa)

# Create an empty raster over RSA at your desired resolution
grid = rast(rsa_vect, resolution = res, crs = crs(rsa_vect))
values(grid) = 1   # fill with dummy values

# Mask everything outside the RSA boundary
grid_masked = mask(grid, rsa_vect)
```

------------------------------------------------------------------------
