---
title: "10-zenodo"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Zenodo basics}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r child-10, child = "_common.Rmd"}
```


```{r setup-10, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "figures/10-",
  out.width = "100%",
  echo = TRUE,
  message = FALSE,
  warning = FALSE
)
```

------------------------------------------------------------------------

# `dissmapr`

## A Novel Framework for Automated Compositional Dissimilarity and Biodiversity Turnover Analysis

------------------------------------------------------------------------

### 1. Deposit all results into Zenodo

All data frames, tables, maps, and standard metadata can be deposited into Zenodo using [`zen4R`](https://github.com/eblondel/zen4R/wiki) or [Zenodo UI](https://zenodo.org/records/10715460).

Below is a minimal `zen4R` workflow showing how to create a new deposition, upload a file, and publish it to Zenodo:

- Make sure `ZENODO_TOKEN` is set in your environment with `deposit:write` (and if plan to publish, `deposit:actions`) scope.   
- `createDeposition()`, `uploadFile()`, and `publishDeposition()` are all methods of the R6 `ZenodoManager` object.   
- Use `downloadFiles()` to pull back any/all files from a given record ID.   

**NOTE**: This will only work if you have an existing Zenodo account.
For step‐by‐step guidance on creating and managing your personal access token, consult the official [Zenodo Developers documentation under Authentication](https://developers.zenodo.org/#authentication).

For a tutorial on using `zen4R` to upload (and download) data, see the official [zen4R vignette](https://cran.r-project.org/web/packages/zen4R/vignettes/zen4R.html).

```{r zenodo}
# # Install and load zen4R if you haven’t already
# # install.packages("zen4R")
# library(zen4R)
# 
# # 1. Authenticate (expects your token in ZENODO_TOKEN)
# token  = Sys.getenv("ZENODO_TOKEN")
# zenodo = ZenodoManager$new(token = token)
# 
# # 2. Create a new (empty) deposition with basic metadata
# dep = zenodo$createDeposition(
#   metadata = list(
#     title       = "My Example Dataset",
#     upload_type = "dataset",
#     description = "A demo upload via zen4R",
#     creators    = list(list(name = "Doe, Jane"))
#   )
# )
# 
# # 3. Upload your local file into that deposition
# zenodo$uploadFile(deposition = dep, file = "path/to/my_data.csv")
# 
# # 4. Publish the deposition (this mints a DOI)
# published = zenodo$publishDeposition(deposition = dep)
# message("Published DOI: ", published$doi)
# 
# # 5. Later on: download all files from that record into a folder
# zenodo$downloadFiles(record_id = dep$id, path = "downloads/")
```

------------------------------------------------------------------------
