---
title: "2-grid"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{2-grid}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup-grid, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "figures/2-",
  out.width = "100%",
  echo = TRUE,
  message = FALSE,
  warning = FALSE
)
```

------------------------------------------------------------------------

# `dissmapr`

## A Novel Framework for Automated Compositional Dissimilarity and Biodiversity Turnover Analysis

------------------------------------------------------------------------

## Spatial aggregation and site-level derivatives

This stage converts point-based occurrence records into a set of harmonised, grid-level data objects that underpin all downstream compositional-dissimilarity analyses. The workflow proceeds as follows:

1. **Generate a reference lattice** with `generate_grid()`. A user-defined cell size (e.g. 0.5°) is draped over the study extent; each cell receives a unique `grid_id`, and point records are assigned to their respective cells without altering their original coordinates.  
2. **Compile per-cell summaries**. For every grid cell the function records observation totals (`obs_sum`), species richness (`spp_rich`), centroid coordinates, and any additional column statistics requested by the user.  
3. **Prepare spatial layers**. The routine returns three coordinated outputs—(i) a raster index (`grid`) for fast raster algebra, (ii) an `sf` polygon layer (`grid_sf`) for cartographic overlays, and (iii) a tidy data-frame (`block_sp`) that links grid attributes to biological information.  
4. **Derive analysis-ready tables**. The grid summary is reshaped into (i) an abundance **site × species matrix** (`site_spp`), (ii) a binary presence–absence version (`sbs`) for incidence metrics, and (iii) a centroid table (`grid_xy`) for rapid spatial joins.  
5. **Map effort and richness**. Observation counts (`obs_cnt`) and species richness (`spp_rich`) are promoted to rasters, providing an immediate visual check on sampling bias and baseline diversity.

The objects created here—`grid_spp`, `site_spp`, `sbs`, `obs_cnt`, `spp_rich`, and their companion rasters—form the spatial backbone for the richness, turnover, and model-based sections that follow.

------------------------------------------------------------------------

### Summarise records by grid centroid using `generate_grid()`

Use `generate_grid` to overlay a user-defined lattice on the study region—whose bounds are inferred from the occurrence data—without modifying the underlying observations themselves. The function constructs a raster‐based grid of the chosen cell size, assigns each point a unique `grid_id`, and compiles a summary of user-specified attributes for every cell. It returns three coordinated outputs: 

*  **grid**: `SpatRaster` with grid index 
*  **grid_sf**: `sf` and `data.frame` i.e. lattice polygon features for mapping or spatial joins 
*  **block_sp**: `data.frame` that records per-cell totals, centroids, and other statistics. 

By aggregating raw records into consistent spatial units, `generate_grid` provides the structured foundation needed for subsequent landscape-scale biodiversity analyses.

------------------------------------------------------------------------

#### Assign records to a grid at a set resolution

Aggregate species records into grid cells of user-specified size (e.g. 0.5°) to enable spatially standardized biodiversity analyses.

```{r load-workspace, echo=FALSE}
load('../inst/savedata/grid_workspace.RData')
```

```{r grid}
grid_list = generate_grid(
  data       = site_spp,
  x_col      = "x",
  y_col      = "y",
  grid_size  = 0.5,
  sum_col_range = 4:ncol(site_spp),
  crs_epsg   = 4326
)

# Check `grid_list` structure
str(grid_list, max.level = 1)

# Optional: Create new objects from list items
aoi_grid = grid_list$grid_sf
grid_spp = grid_list$block_sp

# Check results
dim(aoi_grid)
head(aoi_grid)

dim(grid_spp)
head(grid_spp[,1:6])
```

#### Generate a data frame 'xy' of site centroids

Extract longitude--latitude coordinates and summary metrics for each occupied grid cell.

```{r site-xy}
# Grid centroids with 'gird_id', 'centroid_lon', 'centroid_lat', 'obs_sum' and `spp_rich`
grid_xy = grid_spp[,c(1:3,5:6)]

# Create species observations data.frame
spp_obs = site_obs

# Check results
dim(grid_xy)
head(grid_xy)

dim(spp_obs)
head(spp_obs)
```

#### Generate a map of RSA with occupied grid cells as centroid points

Visualize observation density across South Africa using centroid-based mapping of gridded biodiversity data.

```{r map-aoi}
ggplot() +
  geom_sf(data = aoi_grid, fill = NA, color = "darkgrey", alpha = 0.5) +
  geom_point(data = grid_spp,
             aes(x = centroid_lon, y = centroid_lat,
                 size = sqrt(obs_sum),
                 color = sqrt(obs_sum))) +
  scale_color_viridis_c(option = "turbo") +
  geom_sf(data = rsa, fill = NA, color = "black") +
  theme_minimal() +
  labs(title = "0.5° Grid with Observation Counts",
       x = "Longitude", y = "Latitude")
```

------------------------------------------------------------------------

### Generate site by species matrix - `site_spp`

Create a matrix of species counts per site for use in biodiversity and dissimilarity analyses.

```{r site-spp}
xy = grid_spp[,2:3]
site_spp = grid_spp[,c(1:3,5:ncol(grid_spp))]

# Check results
dim(xy)
head(xy)

dim(site_spp)
head(site_spp[,1:6])
```

#### Fetch record of occurrence counts calculated using `generate_grid()`

`generate_grid` summarised the number of species observations per grid cell to assess sampling effort.

```{r sam-eff}
# Sampling effort = observation counts
sam_eff = grid_spp[, c("grid_id","centroid_lon","centroid_lat","obs_sum")]
obs_cnt = grid_spp[,c(1:3,5)]

# Check results
dim(obs_cnt)
head(obs_cnt)
```

#### Generate a binary (presence/absence) data frame 'sbs'

Convert species abundance data to presence/absence format (`site_spp_pa`) for binary dissimilarity analyses.

```{r sbs}
# Create site-by-species matrix
sbs = site_spp %>%
  mutate(across(all_of(sp_cols), ~ ifelse(. > 0, 1, 0)))

# Check results
dim(sbs)
head(sbs[,1:6])

# Assuming 'sp_cols' is a vector of column names
site_spp_pa = site_spp %>%
  mutate(across(all_of(sp_cols), ~ ifelse(!is.na(.) & . > 0, 1, 0)))

# Check results
dim(site_spp_pa)
head(site_spp_pa[,1:6])
```

#### Fetch species richness values calculated using `generate_grid()`

`generate_grid()` calculated species richness per site based on binary presence/absence records.

```{r spp-rich}
# Species richness
spp_rich = grid_spp[, c("grid_id","centroid_lon","centroid_lat","spp_rich")]

# Check results
dim(site_spp_pa)
head(site_spp_pa[,1:6])
```

#### Generate 'obs_cnt' and 'spp_rich' raster maps

`generate_grid` also creates a `SpatRast` to help visualise sampling effort and species richness over the study area.

```{r eff-rich}
# Species richness and sampling effort SpatRast
ras_effRich = grid_list$grid[[2:3]]

# Plot `ras_effRich` using colour palettes from `viridis`
# plot(sqrt(ras_effRich), col = turbo(100))
plot(
  sqrt(ras_effRich),
  col   = turbo(100),
  colNA = NA,            # <-- NA cells completely transparent
  axes  = FALSE
)

# plot(vect(rsa), add = TRUE, border = "black")
```

***Note**: occurrence coordinates are only used for assigning then into grids; they are not needed beyond this step.*

------------------------------------------------------------------------
