---
title: "Environmental data for sites"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Environmental data for sites}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r setup-3, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "figures/3-",
  out.width = "100%",
  echo = TRUE,
  message = FALSE,
  warning = FALSE
)
```


```{r read_files, include=FALSE}
# At the start of the next vignette, explicitly load files from previous vignette:
inputs = readRDS("_data/2_setup.rds")

rsa = inputs$rsa
grid_sf = inputs$grid_sf
grid_spp = inputs$grid_spp
effRich_r = inputs$effRich_r
grid_spp_pa = inputs$grid_spp_pa
sp_cols = inputs$sp_cols
```


------------------------------------------------------------------------

# `dissmapr`

## A Novel Framework for Automated Compositional Dissimilarity and Biodiversity Turnover Analysis

------------------------------------------------------------------------

### 1. Generate site by environment matrix using `get_enviro_data()`

Spatial models are most informative when each sampling unit couples a biological response (in this example, **sampling effort** and **species richness**) with the same suite of environmental predictors.  
`get_enviro_data()` attaches environmental predictors to each grid cell via a six-stage routine:    
   o **buffer** *the analysis lattice*,   
   o **retrieve** *or read the required rasters*,   
   o **crop** *them to the buffered extent*,   
   o **extract** *raster values at every grid-cell centroid*,   
   o **interpolate** *any missing data gaps*, and   
   o **append** *the finished covariate set to the grid summary*.

The subsections below implement this workflow:

1. **Download and sample 19 WorldClim bioclim variables**: obtains the 5-arc-min (~10 km) [WorldClim v2.1](https://worldclim.org/data/worldclim21.html), returns `bio` stack via [`geodata`](https://github.com/rspatial/geodata), crops it, and attaches climate values to every centroid.
2. **Bind climate, effort, and richness into one raster stack**: combines √-scaled effort (`obs_sum`), √-scaled richness (`spp_rich`), and the 19 climate layers into a single `SpatRast` aligned to the 0.5° grid.
3. **Inspect the extracted covariates**: produces a quick map (e.g. mean annual temperature) and previews the data to verify alignment and plausibility.
4. **Assemble a modelling matrix**: consolidates coordinates, effort, richness, and all climate predictors into a tidy data frame (`grid_env`) ready for statistical modelling.
5. *Optional >>* **Reproject centroids for metric-space analyses**: converts centroid coordinates from `WGS-84` ([EPSG:4326](https://epsg.io/4326)) to a `Albers Equal-Area` projection ([EPSG:9822](https://epsg.io/9822)) when analyses require distances in metres.

---

**Download and sample 19 WorldClim bioclim variables**   
Fetch the 5-arc-min (~10 km) bioclim stack via `geodata` package and attach values to every centroid.

```{r get-enviro}
# Retrieve 19 bioclim layers (≈10-km, WorldClim v2.1) for all grid centroids
data_path = system.file("extdata", package = "dissmapr")               # cache folder for rasters
enviro_list = dissmapr::get_enviro_data(
  data       = grid_spp,                  # centroids + obs_sum + spp_rich
  buffer_km  = 10,                        # pad the AOI slightly
  source     = "geodata",                 # WorldClim/SoilGrids interface
  var        = "bio",                     # bioclim variable set
  res        = 5,                         # 5-arc-min ≈ 10 km
  path       = data_path,
  sp_cols    = 7:ncol(grid_spp),          # ignore species columns
  ext_cols   = c("obs_sum", "spp_rich")   # carry effort & richness through
)

# Quick checks 
str(enviro_list, max.level = 1)

# (Optional) Assign concise layer names for readability
# Find names here https://www.worldclim.org/data/bioclim.html
names_env = c("temp_mean","mdr","iso","temp_sea","temp_max","temp_min",
              "temp_range","temp_wetQ","temp_dryQ","temp_warmQ",
              "temp_coldQ","rain_mean","rain_wet","rain_dry",
              "rain_sea","rain_wetQ","rain_dryQ","rain_warmQ","rain_coldQ")
names(enviro_list$env_rast) = names_env

# (Optional) Promote frequently-used objects
env_r = enviro_list$env_rast    # cropped climate stack
env_df = enviro_list$env_df      # site × environment data-frame

# Quick checks 
env_r
dim(env_df); head(env_df)
```

*`get_enviro_data()` buffered the grid centroids by 10 km, fetched the requested rasters, cropped them, extracted values at each centroid, filled isolated NAs, and merged the results with `obs_sum` and `spp_rich`.*

---

### 2. Bind climate, effort, and richness into one raster stack
  
Fuse the √-scaled sampling‐effort (`obs_sum`) and richness (`spp_rich`) layers with the 19 `bioclim` rasters into a single, co-registered `SpatRast.` A unified stack ensures that all predictors share the same grid, streamlining downstream map algebra, multivariate modelling, and spatial cross-validation.

```{r ras, fig.alt="Environmental Covariates"}
# Resample climate stack to the 0.5 ° grid and concatenate
env_effRich_r = c(
  effRich_r,                               # effort + richness
  resample(env_r, effRich_r)          # aligns 19 climate layers
)

# 2. Open a 2×2 layout and plot each layer + outline
old_par = par(mfrow = c(2, 2), # multi‐figure by row: 1 row and 2 columns 
              mar = c(1, 1, 1, 2))  # margins sizes: bottom (1 lines)|left (1)|top (1)|right (2)

titles = c("Sampling effort (√obs count)",
            "Species richness (√unique count)",
            "BIO1: Annual Mean Temperature", 
            "BIO2: Mean Diurnal Temperature Range")

for (i in 1:4) {
  plot(env_effRich_r[[i]],
       col      = viridisLite::turbo(100),
       colNA    = NA,
       axes     = FALSE,
       main     = titles[i],
       cex.main = 0.8)                        # smaller title
  plot(terra::vect(rsa), add = TRUE, border = "black", lwd = 0.4)
}

par(old_par) 
```

---

### 3. Inspect the extracted covariates

Environmental data were linked to grid centroids using `get_enviro_data()`, now visualise the spatial variation in selected climate variables to check results.

```{r env-df, fig.alt="Grid-cell mean annual temperature (√-scaled)"}
# Make column headers explicit
# names(env_df)[1:5] = c("grid_id","centroid_lon","centroid_lat","obs_sum","spp_rich")

# Simple check of dimensions and first rows
dim(env_df)
head(env_df[, 1:6])

# Quick map of mean annual temperature (√-scaled bubble size)
ggplot() +
  geom_sf(data = grid_sf, fill = NA, colour = "darkgrey", alpha = 0.4) +
  geom_point(data = env_df,
             aes(x = centroid_lon, 
                 y = centroid_lat,
                 colour = bio01),
             shape = 15,
             size = 3) +
  # scale_size_continuous(range = c(2,6)) +
  scale_colour_viridis_c(option = "turbo") +
  geom_sf(data = rsa, fill = NA, colour = "black") +
  theme_minimal() +
  labs(title = "Grid-cell mean annual temperature (√-scaled)",
       x = "Longitude", y = "Latitude")
```

*Goal of this plot is to quickly check that the environmental predictors (e.g. `bio01` >> mean annual temperature) line up with the 0.5° grid.*

---

### 4. Assemble the modelling matrix `grid_env`

Compile a *site × environment* data frame (`grid_env`) in which each 0.5° cell contributes one row containing centroid coordinates, √-scaled sampling effort, species richness, and the 19 `bioclim` predictors. The resulting matrix is immediately usable for GLMs, GAMs, machine-learning, ordination, and β-diversity analyses.

```{r grid-env}
# Build the final site × environment table
grid_env = env_df %>%
  dplyr::select(grid_id, centroid_lon, centroid_lat,
                obs_sum, spp_rich, dplyr::everything())

str(grid_env, max.level = 1)
head(grid_env)
```

---

### 5. Reproject centroids for metric-space analyses** using `sf::st_transform()`[OPTIONAL]

Certain analyses (e.g. spatial clustering, variogram modelling) require coordinates in metres rather than degrees.
The snippet below converts the centroid layer to an `Albers Equal-Area` projection.

```{r reproject}
# Convert the centroid columns to an sf object
centroids_sf = sf::st_as_sf(
  grid_env,
  coords = c("centroid_lon", "centroid_lat"),
  crs    = 4326,          # WGS-84
  remove = FALSE
)

# Reproject to Albers Equal Area (EPSG 9822)
centroids_aea = sf::st_transform(centroids_sf, 9822)

# Append projected X–Y back onto the data-frame
grid_env = cbind(
  grid_env,
  sf::st_coordinates(centroids_aea) |>
    as.data.frame() |>
    setNames(c("x_aea", "y_aea"))   # rename within the pipeline
)
names(grid_env)
head(grid_env[, c("grid_id","centroid_lon","centroid_lat","x_aea","y_aea")])
```

*At this point every grid cell has species metrics, climate predictors, and is optionally projected into metre coordinates, all in a single tidy object.*

---

### 6. Diagnose and mitigate collinearity with `rm_correlated()`

Highly inter-correlated predictors inflate variance, bias coefficient estimates, and complicate ecological inference.  
`rm_correlated()` screens the environmental matrix for pairwise correlations that exceed a user-defined threshold (here |r| > 0.70), then iteratively prunes the variable with the highest average absolute correlation.  The routine

1. Computes a Pearson (default) **Correlation** matrix for the supplied columns;  
2. **Ranks** variables by their mean absolute correlation;  
3. **Discards** the worst offender, recomputes the matrix, and repeats until all remaining pairs lie below the threshold;  
4. *Optional >>* displays the final **Correlation heat-map** for visual QC.

The result is a reduced predictor set that retains maximal information while minimising multicollinearity.

```{r var-vif, message = FALSE, fig.alt="Environmental pairwise correlations"}
# (Optional) Rename BIO
names(env_df) = c("grid_id", "centroid_lon", "centroid_lat", names_env, "obs_sum", "spp_rich")
  
# Run the filter and compare dimensions
# Filter environmental predictors for |r| > 0.70
env_vars_reduced = dissmapr::rm_correlated(
  data       = env_df[, c(4, 6:24)],  # drop ID + coord columns
  cols       = NULL,                  # infer all numeric cols
  threshold  = 0.70,
  plot       = TRUE                   # show heat-map of retained vars
)

# Before vs after
c(original = ncol(env_df[, c(4, 6:24)]),
  reduced  = ncol(env_vars_reduced))
```

*`env_vars_reduced` now contains a decorrelated subset of climate predictors suitable for stable GLMs, GAMs, machine-learning, or ordination workflows.*

---

```{r save-state, include=FALSE}
## Save state for downstream vignettes
# Each vignette runs in a clean R session. We therefore **persist state explicitly**.
dir.create("_data", showWarnings = FALSE, recursive = TRUE)

saveRDS(
  list(
    grid_env = grid_env,
    env_vars_reduced = env_vars_reduced,
    grid_spp_pa = grid_spp_pa,
    sp_cols = sp_cols
  ),
  file = "_data/3_setup.rds",
  compress = "xz"
)

```
