---
title: "7-bioregion"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Map bioregions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r child-7, child = "_common.Rmd"}
```


```{r setup-7, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "figures/7-",
  out.width = "100%",
  echo = TRUE,
  message = FALSE,
  warning = FALSE
)
```


```{r load-change-rasters, include = FALSE}
grid_masked = terra::rast(system.file("extdata", "grid_masked.tif", package = "dissmapr"))
```

------------------------------------------------------------------------

# `dissmapr`

## A Novel Framework for Automated Compositional Dissimilarity and Biodiversity Turnover Analysis

------------------------------------------------------------------------

### 1. Run clustering analyses using `map_bioreg()` to map bioregions

In this step we translate our site‐level ζ₂ predictions into spatial bioregions. Calling `map_bioreg()` on the predictors_df does the following:

1. z-scales the predicted turnover, longitude and latitude;
2. fits four clustering algorithms (k-means, PAM, hierarchical and GMM);
  - **k-means** partitions points around centroids and is fast for large data sets.
  - **PAM** (Partitioning Around Medoids) is a medoid-based analogue of k-means that is more robust to outliers.
  - **Hierarchical** agglomerative clustering builds a dendrogram and then “cuts” it at the chosen k, capturing nested structure in the data.
  - **GMM** (Gaussian Mixture Model) treats clusters as multivariate normal distributions and assigns each point by maximum likelihood.
3. realigns each method’s labels to the k-means solution for consistency;
4. builds both nearest-neighbour and thin-plate-spline interpolated surfaces;
5. returns the raw cluster assignments and gridded rasters, and—because `show_plot=TRUE`—draws a 2×2 panel of maps.

The result is a set of complementary bioregion maps and rasters you can use to compare how different algorithms partition the landscape based on compositional turnover and geography.

```{r zeta-cluster, fig.width=11.25, fig.height=9, warning=FALSE, message=FALSE}
# Run `map_bioreg` function to generate and plot clusters
bioreg_result = map_bioreg(
  data = predictors_df,
  scale_cols = c("pred_zeta", "centroid_lon", "centroid_lat"),
  clus_method = "all", # K-means, PAM, Hierarchical and GMM clustering
  show_plot = TRUE,
  interp = "both",
  x_col ='centroid_lon',
  y_col ='centroid_lat',
  bndy_fc = rsa
)
```

---

### 2. Map future bioregions using `map_bioreg()`

Below we expand our workflow to map the forecasted ζ₂  bioregions under three extreme climate futures (2030, 2040, 2050) alongside the current scenario.
To see how the bioregional partitions shift, we split `all_preds` by scenario and apply `map_bioreg()` (k-means + hierarchical, both NN and TPS interpolation). We then extract the hierarchical cluster layers, mask them to our study area, and plot all four maps in a 2×2 layout:


```{r future-cluster}
# 1) Split your combined predictions by scenario into a named list
by_scn = split(all_preds, all_preds$scenario)

# 2) For each scenario, call map_bioreg() with k‐means + hierarchical, no plots
hier_results = purrr::imap(
  by_scn,
  ~ map_bioreg(
      data        = .x,
      scale_cols  = c("pred_zetaExp", "centroid_lon", "centroid_lat"),
      clus_method = c("kmeans", "hclust"),
      show_plot   = FALSE,
      interp      = "both",            # only NN interpolation to avoid the TPS bug
      x_col       = "centroid_lon",
      y_col       = "centroid_lat"
    )
)

# 3) Give the outer list the scenario names
names(hier_results) = names(by_scn)

# The result is a named list of map_bioreg outputs,
# one element per scenario (current, 2030, 2040, 2050).
str(hier_results, max.level = 1)

# Create SpatRast
future_r = c(hier_results[["current"]]$clusters$hc,
             hier_results[["2030"]]$clusters$hc,
             hier_results[["2040"]]$clusters$hc,
             hier_results[["2050"]]$clusters$hc)
names(future_r)

# 4) Mask `result_bioregDiff` to the RSA boundary
mask_future_r = terra::mask(resample(future_r, grid_masked, method = "near"), grid_masked)

# 5) Quick visual QC in a 3×2 layout
old_par = par(mfrow = c(3, 2), mar = c(1, 1, 1, 5))
titles = c("Current",
            "2030",
            "2040",                       
            "2050")

for (i in 1:4) {
  plot(mask_future_r[[i]],
       # col      = viridisLite::turbo(100),
       col      = viridis(100, direction = -1),
       colNA    = NA,
       axes     = FALSE,
       main     = titles[i],
       cex.main = 0.8)                        # smaller title
  plot(terra::vect(rsa), add = TRUE, border = "black", lwd = 0.4)
}

par(old_par)
```

This end-to-end workflow shows how predicted turnover patterns and resulting bioregions might shift as climate warms and rainfall changes, highlighting potential future reorganization of biodiversity hotspots.

------------------------------------------------------------------------
