---
title: "User-defined grid"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{User-defined grid}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r setup-2, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "figures/2-",
  out.width = "100%",
  echo = TRUE,
  message = FALSE,
  warning = FALSE
)
```

```{r read_files, include=FALSE}
# Load libraries
library(ggplot2)

# Load the objects this article needs from the single bundled snapshot.
inputs = readRDS(system.file("extdata", "dissmapr_vignettes.rds", package = "dissmapr"))

site_obs = inputs$site_obs
site_spp = inputs$site_spp
n_sp = inputs$n_sp
sp_cols = inputs$sp_cols
```

# `dissmapr`

## A Novel Framework for Automated Compositional Dissimilarity and Biodiversity Turnover Analysis

### 1. User-defined area of interest and grid resolution

Defining the geographic extent and an analysis grid early ensures that all subsequent data extraction, aggregation, and visualisation tasks are carried out within a consistent spatial framework. In this vignette we:

1. **Load the national boundary of South Africa** to set our area of interest (AoI).
2. **Select a working resolution** of 0.5° (≈ 55 km) to balance spatial detail with computational cost.
3. **Convert the AoI to a `terra` vector** so that raster operations run efficiently.
4. **Create a blank raster template** using the chosen resolution and the AoI’s CRS (Coordinate Reference System).
5. **Populate the raster with placeholder values** (here simply 1).
6. **Mask the raster to the AoI** so that only cells whose centroids fall within South Africa remain.

```{r aoi}
# 1. Load the national boundary 
# The shapefile is shipped with the package for full reproducibility.
rsa = sf::st_read(system.file("extdata", "rsa.shp", package = "dissmapr"))

# 2. Choose a working resolution 
# A 0.5‑degree cell size strikes a balance between computational load and
# the spatial resolution at which national‑level biodiversity patterns remain
# interpretable.
res = 0.5   # decimal degrees° (≈ 55 km at the equator)

# 3. Convert the AoI to a 'terra' vector 
# 'terra' supports fast raster operations; converting now avoids repeated
# coercion later.
rsa_vect = terra::vect(rsa)

# 4. Initialise a blank raster template 
# The template inherits the AoI’s coordinate reference system (CRS) and is
# discretised into equally‑sized cells according to the resolution chosen.
grid = terra::rast(rsa_vect, resolution = res, crs = terra::crs(rsa_vect))

# 5. Populate the raster with placeholder values 
# We simply assign the value 1 to every cell; the values themselves are
# irrelevant at this stage—the grid’s geometry is what matters.
terra::values(grid) = 1

# 6. Clip the raster to the AoI 
# Any cells whose centroids fall outside the boundary are set to NA, thereby
# restricting subsequent computations to the AoI only.
grid_masked = terra::mask(grid, rsa_vect)
# grid_masked is now a 0.5° lattice clipped to South Africa and will serve as the common spatial denominator for all downstream summaries.
```

### 2. Summarise records by grid centroid using `generate_grid()`

With the national lattice in place, we can now **condense point-level observations to grid cells** using `generate_grid()` to:

1. **Construct a bounding grid**: Expands the extent of input points and tessellates it with square cells of the chosen size (here 0.5°).
2. **Allocate a `grid_id`**: Every record inherits the ID of the cell in which it falls.
3. **Aggregate user-selected columns** within each occupied cell, returning:  
   - `grid_spp`: species counts / abundances.  
   - `grid_spp_pa`: the same matrix recoded to presence (1) / absence (0) for binary dissimilarity metrics.  
   - `obs_sum`: total observations across the aggregated columns.  
   - `spp_rich`: number of columns with a non-zero count (simple species richness).
4. **Compute cell centroids** and optional assign mapsheet codes (useful for atlasing projects).
5. **Rasterise key layers** (`grid_id`, `obs_sum`, `spp_rich`) for fast map algebra.
6. **Return four spatial objects** ready for further analysis:  
   - `grid_r`: multi-layer `SpatRaster`  
   - `grid_sf`: polygon lattice with centroids & summaries  
   - `grid_spp`: abundance table (per cell × species)  
   - `grid_spp_pa`: binary presence/absence table (same dimensions as `grid_spp`)

Because every observation is now referenced to a regular grid, all downstream statistics and graphics are standardised to the same sample area.

```{r grid}
# Generate a 0.5° grid summary for the point dataset `site_spp`
grid_list = dissmapr::generate_grid(
  data          = site_spp,           # point data with x/y + species columns
  x_col         = "x",                # longitude column
  y_col         = "y",                # latitude  column
  grid_size     = 0.5,                # cell size in degrees
  sum_cols      = 4:ncol(site_spp),   # columns to aggregate
  crs_epsg      = 4326                # WGS84
)

# Inspect the returned list 
str(grid_list, max.level = 1)

# (Optional) Promote list items to named objects 
grid_r = grid_list$grid_r   # raster summary
grid_sf = grid_list$grid_sf   # polygons for mapping or joins
grid_spp = grid_list$grid_spp # tabular summary per cell
grid_spp_pa = grid_list$grid_spp_pa # presence/absence summary

# Quick checks 
dim(grid_sf); head(grid_sf)
dim(grid_spp); head(grid_spp[, 1:8])
dim(grid_spp_pa); head(grid_spp_pa[, 1:8])
```

`grid_spp` now serves as the **site‑level backbone** for modelling (e.g. spatial GLMs) or visualisation (e.g. dot plots), whereas `grid_spp_pa` slots directly into Jaccard- or Sørensen-based beta-diversity workflows. `site_spp` retains the raw observation detail for drill‑down analyses.

### 3. Visualise observation density across South Africa

With the grid summaries in hand we can now **map the spatial distribution of observation effort**. The recipe below layers three geometric objects in a single `ggplot2` call:

1. **Grid polygons (`grid_sf`)**: Outlined in semi‑transparent grey to give a subtle sense of the analytical lattice without overwhelming the figure.
2. **Centroid points (`grid_spp`)**: Plotted using longitude/latitude coordinates and symbol attributes that encode sampling intensity. For example, below **size & colour** are mapped to `sqrt(obs_sum)`.  We use `sqrt()` because a square‑root transform is often preferable when counts span large orders of magnitude as it compresses large values while still highlighting structure among sparsely sampled cells.
3. **National border (`rsa`)**: Emphasised in solid black to anchor the map in a familiar outline.

A perceptually uniform `Viridis` palette (`option = "turbo"`) supports colour‑blind accessibility, while `theme_minimal()` removes visual clutter so the data can speak for themselves.

```{r map-aoi, fig.alt="Observation density across South Africa (0.5° grid)"}
ggplot2::ggplot() +
  # 1. grid polygons as subtle backdrop 
  geom_sf(data = grid_sf, fill = NA, colour = "darkgrey", linewidth = 0.2, alpha = 0.5) +
  
  # 2. centroids sized/coloured by sampling effort 
  geom_point(
    data = grid_spp,
    aes(x = centroid_lon, y = centroid_lat,
        size  = sqrt(obs_sum),
        colour = sqrt(obs_sum)),
    alpha = 0.8
  ) +
  
  # Divergent colour scale 
  scale_colour_viridis_c(option = "turbo", name = "√ Observations") +
  scale_size_continuous(name = "√ Observations", guide = "none") +
  
  # 3. national outline 
  geom_sf(data = rsa, fill = NA, colour = "black", linewidth = 0.4) +
  
  theme_minimal() +
  labs(
    title = "Observation density across South Africa (0.5° grid)",
    x = "Longitude", y = "Latitude"
  )
```

---

### 4. Visualise sampling effort and richness

`generate_grid()` also returns a three-layer `SpatRaster` (`grid_r`) whose second and third bands store cell-level metrics:

- `obs_sum`: Total observations aggregated across the chosen species columns (units = observation count)
- `spp_rich`:	Number of species (non-zero columns) recorded in the cell	(units = unique species count)

The chunk below extracts those two layers, applies a square-root stretch (to dampen the influence of very large counts), and renders them side-by-side with a perceptually uniform turbo palette.

```{r eff-rich, fig.alt="Sampling effort (√obs count) and Species richness (√unique count)"}
# 1. Extract & transform layers (use terra method explicitly)
effRich_r = sqrt(grid_r[[c("obs_sum", "spp_rich")]])

# 2. Save and reset graphics state safely
old_par = par(no.readonly = TRUE)
on.exit(par(old_par), add = TRUE)

layout(matrix(1:2, nrow = 1))

titles = c(
  "Sampling effort (√obs count)",
  "Species richness (√unique count)"
)

for (i in 1:2) {
  terra::plot(
    effRich_r[[i]],
    col    = viridisLite::turbo(100),
    breaks = 100,
    colNA  = NA_character_,
    axes   = FALSE,
    main   = titles[i],
    cex.main = 0.8
  )
  terra::plot(terra::vect(rsa), add = TRUE, border = "black", lwd = 0.4)
}
```

These maps quickly reveal where sampling effort is concentrated and how species richness varies across the landscape—useful diagnostics before any downstream modelling.

```{r session-info}
sessionInfo()
```
