---
title: "Predict community turnover"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Predict community turnover}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup-6, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "figures/6-",
  out.width = "100%",
  echo = TRUE,
  message = FALSE,
  warning = FALSE
)
```

# `dissmapr`

## Predicting Spatial Patterns of Biodiversity Turnover

This vignette focuses on generating spatial predictions from fitted biodiversity-turnover models. The goal is to show how model outputs can be projected across the analysis grid to create continuous spatial layers for mapping and interpretation.

To keep the example reproducible and quick to run, we use a small set of example objects bundled with `dissmapr`. The setup chunk below loads the required packages, reads the bundled data snapshot, and unpacks the prediction outputs, raster template, species data, and boundary objects needed for spatial prediction.

```{r read_files}
# Load the packages used in this vignette.
library(dissmapr)
library(ggplot2)
library(dplyr)
library(purrr)

# Load the bundled example data snapshot.
inputs = readRDS(system.file("extdata", "dissmapr_vignettes.rds", package = "dissmapr"))

# Unpack the example objects used below.
grid_spp_pa = inputs$grid_spp_pa           # Presence-absence species data
env_vars_reduced = inputs$env_vars_reduced # Selected environmental variables
grid_env = inputs$grid_env                 # Grid-level environmental data
sp_cols = inputs$sp_cols                   # Species column names
zeta2 = inputs$zeta2                       # Pairwise zeta-diversity output
ispline_gdm_tab = inputs$ispline_gdm_tab   # Fitted I-spline/GDM results table
grid_spp = inputs$grid_spp                 # Grid-level species data
rsa = inputs$rsa                           # South Africa boundary
```

### 1. Predict current Zeta Diversity (zeta2) using `predict_dissim()`

In this step we use our fitted order-2 GDM (`zeta2`) to generate a spatial map of pairwise compositional turnover (ζ₂) under current conditions. By `calling predict_dissim()` we

1. compute each site’s sampling‐effort‐adjusted species richness and mean distance to all other sites;   
2. apply the same environmental I-spline transformations used in the model;   
3. predict the Jaccard-scaled turnover (ζ₂) on the 0–1 scale;   
4. optionally plot the resulting heatmap with your study boundary overlaid.

We set a random seed for reproducibility (so Monte Carlo sampling inside `predict_dissim()` yields the same results each time), pull out just the species columns once, then inspect the returned `predictors_df` to confirm its dimensions, column names, and a quick peek at the key model outputs.

```{r predict-zeta2, fig.alt="Predicted pairwise compositional turnover"}
# Predict current zeta diversity using `predict_dissim` with sampling effort, geographic distance and environmental variables
# Only non-colinear environmental variables used in `zeta2` model
set.seed(123) # set.seed to generate exactly the same random results i.e. sam=100
spp_cols = names(grid_spp_pa[,-(1:6)])
predictors_df   = dissmapr::predict_dissim(
  grid_spp      = grid_spp_pa,
  species_cols  = spp_cols,
  env_vars      = env_vars_reduced,# env_vars_reduced[,-8]
  zeta_model    = zeta2, # From simple order 2 run above
  # zeta_model = ispline_gdm_tab$zeta_gdm_list[[1]], # From list of Zeta.msgdm models
  grid_xy       = grid_env,
  x_col         = "centroid_lon",
  y_col         = "centroid_lat",
  bndy_fc       = rsa, # Optional feature collection to plot as boundary
  show_plot     = TRUE
)

# Check results
dim(predictors_df)
names(predictors_df)
head(predictors_df[,5:11])
```

### 2. Predict future Zeta Diversity using `predict_dissim()`

Below we expand our workflow to forecast how ζ₂ respond to three extreme climate futures (2030, 2040, 2050) alongside the current scenario.

First, we define and center-scale each future by adding large temperature increments and rainfall multipliers, then bundle them into a named list of four environmental data frames:
   
```{r sim-future, message=FALSE}
# 1. Identify species & env columns
# spp_cols  = names(grid_spp_pa)[-(1:6)]
all_vars  = names(env_vars_reduced)
temp_vars = grep("^temp", all_vars, value = TRUE)
rain_vars = grep("^rain", all_vars, value = TRUE)
iso_vars = grep("^iso", all_vars, value = TRUE)
obs_var = "obs_sum"

# 2. Extreme future shifts
horizons     = c("2030","2040","2050")
mean_delta   = list(
  temp  = c("2030"= +2,  "2040"= +4,  "2050"= +6),
  iso   = c("2030"= +0.5,"2040"= +1.0,"2050"= +1.5),
  rain  = c("2030"= 0.9, "2040"= 0.8, "2050"= 0.7),
  effort= c("2030"= 1.3, "2040"= 1.6, "2050"= 2.0)
)
exagg_factor = c("2030"=1.5, "2040"=2.0, "2050"=2.5)

# 3) helper to amplify deviation from the mean
amplify = function(x, factor) {
  m = mean(x, na.rm=TRUE)
  m + (x - m) * factor
}

# # 4. Save original scaling parameters
# sc_params = scale(env_vars_reduced)
# mu    = attr(sc_params, "scaled:center")
# sigma = attr(sc_params, "scaled:scale")

# 4. Build list of future env tibbles
env_scenarios = c(
  list(current = env_vars_reduced),
  map(horizons, function(yr) {
    df = env_vars_reduced
    # mean shifts
    df[temp_vars] = df[temp_vars] + mean_delta$temp[yr]
    df[iso_vars]  = df[iso_vars]  + mean_delta$iso[yr]
    df[rain_vars] = df[rain_vars] * mean_delta$rain[yr]
    df[[obs_var]] = df[[obs_var]] * mean_delta$effort[yr]
    # amplify spatial variation
    df[temp_vars] = map_dfc(df[temp_vars], amplify, factor = exagg_factor[yr])
    df[iso_vars]  = map_dfc(df[iso_vars],  amplify, factor = exagg_factor[yr])
    # clamp obs_sum
    df[[obs_var]] = pmin(pmax(df[[obs_var]], 50), 8000)
    df
  }) |> set_names(horizons)
)
# names(env_scenarios) = names(horizons)

# 5. Prepend current conditions
# env_scenarios = c(list(current = env_vars_reduced), env_futures)
str(env_scenarios, max.level = 1)
```

Next, we loop through each scenario, re-apply the original centering and scaling, and call our `predict_dissim()` helper to compute ζ₂. We tag each result with its scenario name and combine them into one tidy data frame:

```{r predict-future, message=FALSE}
set.seed(123)

scenario_dfs = purrr::imap(env_scenarios, ~ {
  df = dissmapr::predict_dissim(
    grid_spp     = grid_spp,
    species_cols = spp_cols,
    env_vars     = .x,
    zeta_model   = zeta2,
    grid_xy      = grid_env,
    x_col        = "centroid_lon",
    y_col        = "centroid_lat",
    skip_scale   = FALSE,
    show_plot    = FALSE
  )
  df$scenario = .y
  df
})
str(scenario_dfs, max.level=1)

# all_preds = bind_rows(scenario_dfs) %>%
#   mutate(scenario = factor(scenario, levels = c("current", names(temp_shifts))))
all_preds = bind_rows(scenario_dfs) 
head(all_preds)
```

We can then quickly compare the spatial ζ₂ surfaces under each future:

```{r predict-future-plot, message=FALSE, fig.alt="Predicted pairwise compositional turnover with climate change"}
ggplot(all_preds,
       aes(centroid_lon, centroid_lat, fill = pred_zetaExp)) +
  geom_tile() +
  facet_wrap(~ scenario, ncol = 2) +
  scale_fill_viridis_c(direction = -1, name = expression(zeta[2])) +
  geom_sf(data = rsa, fill = NA, color = "black", inherit.aes = FALSE) +
  coord_sf() +
  labs(x = "Longitude", y = "Latitude",
       title = expression("Predicted ζ"[2] * " under current & future scenarios")) +
  theme_minimal() +
  theme(strip.text = element_text(face = "bold"),
        panel.grid = element_blank())
```

```{r session-info}
sessionInfo()
```
