---
title: "Map bioregions"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Map bioregions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup-7, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "figures/7-",
  out.width = "100%",
  echo = TRUE,
  message = FALSE,
  warning = FALSE
)
```

```{r read_files}
# Load the objects this article needs from the single bundled snapshot.
library(dissmapr)
library(terra)
library(RColorBrewer)
inputs = readRDS(system.file("extdata", "dissmapr_vignettes.rds", package = "dissmapr"))

predictors_df = inputs$predictors_df
all_preds = inputs$all_preds
grid_masked = terra::mask(terra::setValues(terra::rast(system.file("extdata", "grid_r.tif", package = "dissmapr"))[[1]], 1), terra::vect(inputs$rsa))
rsa = inputs$rsa
grid_spp = inputs$grid_spp
sp_cols = inputs$sp_cols

future_nn = terra::rast(system.file("extdata", "future_nn.tif", package = "dissmapr"))
```

# `dissmapr`

## A Novel Framework for Automated Compositional Dissimilarity and Biodiversity Turnover Analysis

### 1. Run clustering analyses using `map_bioreg()` to map bioregions

In this step we translate our site‐level ζ₂ predictions into spatial bioregions. Calling `map_bioreg()` on the predictors_df does the following:

1. z-scales the predicted turnover, longitude and latitude;
2. fits four clustering algorithms (k-means, PAM, hierarchical and GMM);
  - **k-means** partitions points around centroids and is fast for large data sets.
  - **PAM** (Partitioning Around Medoids) is a medoid-based analogue of k-means that is more robust to outliers.
  - **Hierarchical** agglomerative clustering builds a dendrogram and then “cuts” it at the chosen k, capturing nested structure in the data.
  - **GMM** (Gaussian Mixture Model) treats clusters as multivariate normal distributions and assigns each point by maximum likelihood.
3. realigns each method’s labels to the k-means solution for consistency;
4. builds both nearest-neighbour and thin-plate-spline interpolated surfaces;
5. returns the raw cluster assignments and gridded rasters, and—because `show_plot=TRUE`—draws a 2×2 panel of maps.

The result is a set of complementary bioregion maps and rasters you can use to compare how different algorithms partition the landscape based on compositional turnover and geography.

```{r zeta-cluster, fig.width=11.25, fig.height=9, warning=FALSE, message=FALSE, fig.alt="Bioregion clusters"}
# Add this to {, fig.width=11.25, fig.height=9, warning=FALSE, message=FALSE}
# Run `map_bioreg` function to generate and plot clusters
bioreg_current = dissmapr::map_bioreg(
  data = predictors_df,
  scale_cols = c("pred_zetaExp", "centroid_lon", "centroid_lat"),
  method = 'all', # Options: c("kmeans","pam","hclust","gmm","all"),
  k_override  = 8,
  interpolate = 'nn', # Options: c("none","nn","tps","all"),
  x_col ='centroid_lon',
  y_col ='centroid_lat',
  res = 0.5, 
  crs = "EPSG:4326",
  plot = TRUE,
  bndy_fc = rsa)

# Check results
str(bioreg_current, max.level=1)
```

### 2. Map future bioregions using `map_bioreg()`

Below we expand our workflow to map the forecasted ζ₂  bioregions under three extreme climate futures (2030, 2040, 2050) alongside the current scenario.
To see how the bioregional partitions shift, we split `all_preds` by scenario and apply `map_bioreg()` (k-means + hierarchical, both NN and TPS interpolation). We then extract the hierarchical cluster layers, mask them to our study area, and plot all four maps in a 2×2 layout:

```{r future-cluster, fig.width=9, fig.height=8, fig.alt="Future bioregion clusters"}
# Split your combined predictions by scenario into a named list
by_scn = split(all_preds, all_preds$scenario)

# For each scenario, call map_bioreg() with all algorithms
bioreg_future = dissmapr::map_bioreg(
  data = by_scn,
  scale_cols = c("pred_zetaExp", "centroid_lon", "centroid_lat"),
  method = 'all', # Options: c("kmeans","pam","hclust","gmm","all"),
  k_override  = 8,
  interpolate = 'nn', # Options: c("none","nn","tps","all"),
  x_col ='centroid_lon',
  y_col ='centroid_lat',
  res = 0.5, 
  crs = "EPSG:4326",
  plot = FALSE,
  bndy_fc = rsa)

# Check results
str(bioreg_future, max.level=1)

# Plot the consensus bioregion clusters per scenario with a single, compact
# shared legend. Bioregions are categorical, so we keep a discrete key (Set3)
# rather than a continuous colour bar.
library(ggplot2)

clust_tbl = bioreg_future$table

# Order scenarios with "current" first, if present
scn_lev = unique(clust_tbl$scenario)
scn_lev = c(intersect("current", scn_lev), setdiff(scn_lev, "current"))
clust_tbl$scenario = factor(clust_tbl$scenario, levels = scn_lev)

ggplot() +
  geom_tile(data = clust_tbl,
            aes(x = centroid_lon, y = centroid_lat,
                fill = factor(cluster_mode))) +
  geom_sf(data = rsa, fill = NA, colour = "black", linewidth = 0.3) +
  facet_wrap(~ scenario) +
  scale_fill_brewer(palette = "Set3", name = "Bioregion", drop = FALSE) +
  theme_minimal(base_size = 9) +
  theme(legend.position = "bottom") +
  guides(fill = guide_legend(nrow = 1)) +
  labs(x = "Longitude", y = "Latitude")
```

Below we visualise the nearest-neighbour interpolated future‐scenario cluster outputs. First, we list the structure of the `bioreg_future` result to confirm available components. We then combine the k-means nearest-neighbour rasters for “current” and each future year into a single `SpatRaster` stack (`future_nn`), and resample, then mask it to the RSA boundary (`mask_future_nn`). Finally, we lay out a 2×2 plot grid, compute a discrete colour palette for each layer based on its unique classes, and render each masked layer with its boundary overlay for a quick inspection of bioregion changes across time.

```{r future-plots, fig.width=9, fig.height=6.5, fig.alt="Smoothed future bioregion clusters"}
# Check results
str(bioreg_future, max.level=1)

# Create SpatRast
# future_nn = c(bioreg_future$nn$current$kmeans_current,
#              bioreg_future$nn$`2030`$kmeans_2030,
#              bioreg_future$nn$`2040`$kmeans_2040,
#              bioreg_future$nn$`2050`$kmeans_2050)
names(future_nn)

# 4) Mask `result_bioregDiff` to the RSA boundary
mask_future_nn = terra::mask(resample(future_nn, grid_masked, method = "modal"), grid_masked)

# 5) Quick visual QC in a 2×2 layout
old_par = par(mfrow = c(2, 2), mar = c(1, 1, 1, 5))
titles = c("Current",
            "2030",
            "2040",
            "2050")

for (i in 1:4) {
  ## 1. how many distinct classes in this layer?
  cls  = sort(unique(values(mask_future_nn[[i]])))
  cls  = cls[!is.na(cls)]
  n    = length(cls)

  ## 2. build a discrete palette of n colours
  pal = if (n <= 12) {
           RColorBrewer::brewer.pal(n, "Set3")                      # native Set3
         } else {
           colorRampPalette(brewer.pal(12, "Set3"))(n) # extended Set3
         }

  ## 3. plot
  plot(mask_future_nn[[i]],
       col      = pal,
       type     = "classes",          # treats values as categories
       colNA    = NA,
       axes     = FALSE,
       legend   = TRUE,
       main     = titles[i],
       cex.main = 0.8)

  plot(terra::vect(rsa), add = TRUE, border = "black", lwd = .4)
}

par(old_par)
```

This end-to-end workflow shows how predicted turnover patterns and resulting bioregions might shift as climate warms and rainfall changes, highlighting potential future reorganization of biodiversity hotspots.

```{r session-info}
sessionInfo()
```
